# Budgeting

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

Most people hate budgeting because they're approaching it with some bad assumptions.  Traditional budgets fail because they assume human behavior is robotic: tracking every penny, restricting every joy, and demanding perfection. The frameworks (drawn from Sethi, Housel, Bogle, and Felix) are designed for how we actually behave.

> "A budget tells your money where to go instead of wondering where it went." (Ramit Sethi)

## The Budget Mindset Shift

Budgets aren't about restriction; they're about *conscious allocation*.  A good budget builds in flexibility, not rigidity. Morgan Housel's insight is powerful here: "*room for error is the most underrated financial concept.*"

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Budget Mindset'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    OldBudget([Old Budget Mindset]) --> Restrict[Restrict Everything]
    Restrict --> Fail(Fails Within Weeks)
    
    NewBudget([Conscious Spending Plan]) --> Prioritize[Prioritize What Matters]
    Prioritize --> Sustain(Sustains for Years)
```

## Fixed vs. Variable Expenses

The first skill in budgeting is correctly categorizing your money. Most people lump everything together, then get overwhelmed by how to allocate or account for the money they are spending.

```{mermaid}
%%| fig-cap: 'Fixed vs. Variable Expenses'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart LR
    Expenses(["Monthly<br>Expenses"]) --> Fixed["FIXED EXPENSES:<br/>Predictable & Recurring"]
    Expenses --> Variable["VARIABLE EXPENSES:<br/>Fluctuate Month to Month"]
    Expenses --> Periodic["PERIODIC EXPENSES:<br/>Annual or Irregular"]
    
    Fixed --> Rent("Rent/Mortgage")
    Fixed --> Insurance("Insurance<br>Premiums")
    Fixed --> Subs("Subscriptions")
    Fixed --> Loans("Loan<br>Payments")
    
    Variable --> Groceries("Groceries")
    Variable --> Utilities("Utilities")
    Variable --> Gas("Gas/Transport")
    Variable --> Dining("Dining<br>Out")
    
    Periodic --> Taxes("Property<br>Taxes")
    Periodic --> CarMaint("Car<br>Maintenance")
    Periodic --> Gifts("Gifts/Holidays")
    Periodic --> Medical("Medical Costs")
```

### Why Categorizing Expenses Matters {.unnumbered}

Fixed and variable expenses should employ different strategies. Periodic expenses are trickier because they are specific to the item/person. 

| Category | Strategy |
|----------|----------|
| **Fixed Expenses** | Lock these in and forget them. The goal is to **minimize them once**, then automate. Biggest leverage point: housing. |
| **Variable Expenses** | Where most "budgeting" effort fails. Use **categories with limits**, not line-by-line tracking. |
| **Periodic Expenses** | The silent budget-killer. Use a **sinking fund**: set aside 1/12 of the annual cost each month. |

### Sinking Fund: Periodic Expenses {.unnumbered}

If your car insurance is \$1,200/year, don't get blindsided every six months. Save \$100/month into a "Car Insurance" sub-account. When the bill arrives, the money is already there. Repeat for every periodic expense:

- Christmas/gifts: \$600/year → \$50/month
- Car maintenance: \$1,200/year → \$100/month
- Annual vacation: \$2,400/year → \$200/month

This single technique eliminates 80% of "budget emergencies."

## Zero-Based vs. Percentage Budgets

Two dominant frameworks exist. Both work; pick the one that matches your personality. Zero-based budgets offer tight control and are  detail-oriented, making them ideal for anyone rebuilding from debt, or for irregular incomes. Percentage budgets are for big-picture thinkers, stable income earners, and people who hate detailed tracking.

### Zero-Based Budgeting {.unnumbered}

Every dollar gets a job. Income minus all allocations equals zero. 

```{mermaid}
%%| fig-cap: 'Zero-Based Budgeting'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart LR
    Income["$5,000 Income"] --> Rent2["$1,500 Rent"]
    Income --> Food[$500 Groceries]
    Income --> Transport[$300 Transport]
    Income --> Invest[$1,000 Investments]
    Income --> Save[$500 Savings]
    Income --> Fun[$700 Fun Money]
    Income --> Misc[$500 Other]
    Misc --> Zero["Total = $5,000<br/>Remaining = $0"]
```

* **Pros:** Maximum awareness, every dollar intentional, exposes leaks immediately.

* **Cons:** High maintenance, can feel restrictive, requires re-budgeting monthly.

### Percentage-Based Budgeting {.unnumbered}

Allocate by percentages, not dollars. This is the foundation of Sethi's Conscious Spending Plan and similar frameworks (50/30/20, etc.).

| Bucket | Conscious Spending Plan (Sethi) | 50/30/20 Rule |
|--------|--------------------------------|---------------|
| Fixed Costs | 50–60% | 50% Needs |
| Investments | 10%+ | 20% Savings/Debt |
| Savings Goals | 5–10% | (combined above) |
| Guilt-Free Spending | 20–35% | 30% Wants |

* **Pros:** Simple, sustainable, scales with income, low maintenance.

* **Cons:** Less granular, can mask category overspending, requires honesty about category boundaries.

### Which Should You Choose? {.unnumbered}

```{mermaid}
%%| fig-cap: 'Which Budget To Choose?'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Choose[Which Budget Style?] --> Question1{In Debt or<br/>Just Starting?}
    Question1 -->|Yes| ZeroBased[Zero-Based<br/>for 6-12 months]
    Question1 -->|No| Question2{Love Spreadsheets<br/>& Details?}
    Question2 -->|Yes| Hybrid[Hybrid Approach]
    Question2 -->|No| Percentage[Percentage-Based<br/>Conscious Spending]
    ZeroBased --> Eventually[Graduate to Percentage<br/>Once Stable]
```

```{r}
#| label: my-rec-budget
#| echo: false
#| results: asis
#| eval: true
co_box(color = "g", 
  header = "My recommendation", 
  contents = "Start with zero-based for 2-3 months to truly understand your spending. Then transition to a percentage-based system with automation. You get the awareness without the long-term burden.")
```

## Build a Budgeting Tools

You can build your own budget using a spreadsheet or code (like the examples below). Applications like Microsoft Excel or Google Sheets have tools and templates for customizing each sheet/tab to fit your needs (search "budget template"). Most personal finance blogs also provide a classic envelope spreadsheet free to download.  

### The Spreadsheet Approach {.unnumbered}

Bogle would approve: **simplicity wins.** A well-built spreadsheet beats most apps because it forces you to engage with your numbers. Here's a basic structure:

```{mermaid}
%%| fig-cap: 'Spreadsheet'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart LR
    Spreadsheet(["Budget<br>Spreadsheet"]) --> Tab1[Tab 1: Income]
    Spreadsheet --> Tab2[Tab 2: Fixed Expenses]
    Spreadsheet --> Tab3[Tab 3: Variable Expenses]
    Spreadsheet --> Tab4[Tab 4: Sinking Funds]
    Spreadsheet --> Tab5[Tab 5: Net Worth Tracker]
    Spreadsheet --> Tab6[Tab 6: Summary Dashboard]
    
    Tab6 --> Metrics("Key Metrics:<br/>Savings Rate<br/>Fixed Cost %<br/>Monthly Cash Flow")
```

### What Actually Matters {.unnumbered}

Don't overthink this. The best budgeting tool is **the one you'll actually use.** 

Features that matter:

1. **Automatic transaction imports** (saves hours per month)
2. **Easy categorization** (you'll do this weekly)
3. **A clear summary view** (your dashboard)
4. **Goal tracking** (motivation matters)

Features that don't matter as much: fancy graphs, AI insights, social features. Substance over style.

## Budgeting Math

Here's the secret most personal finance content avoids: *budgeting requires only middle-school math.* However, you need to be fluent in a few specific calculations.

Each calculation below is also written in R and Python. Because you'll re-use these formulas constantly, we'll wrap each one in a **function**: a named, reusable formula you define once and call with whatever numbers you have. In R, the pattern is `name <- function(inputs) { ... }`; in Python, it's `def name(inputs):`. Anything after a `#` is a comment the computer ignores. The R code uses the [tidyverse](https://www.tidyverse.org/) (loaded once below) and the Python code uses [`numpy`](https://numpy.org/) where vectors help.

```{r}
#| label: math-libs-r
#| message: false
library(tidyverse)
```

```{python}
#| label: math-libs-py
import numpy as np
```

### Percentages (Your Most-Used Skill) {.unnumbered}

You need to instantly answer: "What percentage of my income is X?"

**Formula:** (Expense ÷ Income) × 100

Example: Rent is $1,500, income is $5,000. → 1,500 ÷ 5,000 = 0.30 → **30% of income on rent.**

Practice until you can do this in your head for the big categories.

::: {.panel-tabset}

## R

```{r}
#| label: percent-r
percent_of_income <- function(expense, income) {
  (expense / income) * 100
}

# rent of $1,500 against $5,000 take-home pay
percent_of_income(expense = 1500, income = 5000)
```

## Python

```{python}
#| label: percent-py
def percent_of_income(expense, income):
    return (expense / income) * 100

# rent of $1,500 against $5,000 take-home pay
percent_of_income(expense=1500, income=5000)
```

:::

### Annualizing and Monthly-izing {.unnumbered}

The skill of converting between time periods. Budgets fail because people don't see annual costs in monthly terms (or vice versa).

**Annual → Monthly:** Divide by 12

**Monthly → Annual:** Multiply by 12

**Weekly → Monthly:** Multiply by 4.33 (not 4)

**Daily habit → Annual:** Multiply by 365

Example: A $5 daily coffee → $5 × 365 = **$1,825/year.** Suddenly it's a real number you can evaluate.

::: {.panel-tabset}

## R

```{r}
#| label: annualize-r
to_annual <- function(amount, period = c("daily", "weekly", "monthly")) {
  period  <- match.arg(period)
  factors <- c(daily = 365, weekly = 52, monthly = 12)
  amount * factors[[period]]
}

to_monthly <- function(amount, period = c("daily", "weekly", "annual")) {
  period  <- match.arg(period)
  factors <- c(daily = 30.44, weekly = 4.33, annual = 1 / 12)
  amount * factors[[period]]
}

to_annual(amount = 5,    period = "daily")    # $5/day coffee → annual
to_monthly(amount = 150, period = "weekly")   # $150/week groceries → monthly
```

## Python

```{python}
#| label: annualize-py
def to_annual(amount, period="daily"):
    factors = {"daily": 365, "weekly": 52, "monthly": 12}
    return amount * factors[period]

def to_monthly(amount, period="daily"):
    factors = {"daily": 30.44, "weekly": 4.33, "annual": 1 / 12}
    return amount * factors[period]

print(to_annual(amount=5,    period="daily"))    # $5/day coffee → annual
print(to_monthly(amount=150, period="weekly"))   # $150/week groceries → monthly
```

:::

### Compound Interest (The Wealth Equation) {.unnumbered}

You don't need to calculate this by hand, but you need to *understand* it.

**Formula:** FV = PV × (1 + r)^n

Where: FV = future value, PV = present value, r = annual return, n = years

**The shortcut you should memorize is the Rule of 72:**

> Years to double your money = 72 ÷ annual return %

At 7% return, money doubles every ~10 years. At 10%, every ~7 years. This single insight reframes every spending decision.

In code, the `^` symbol (R) and `**` symbol (Python) both mean "raise to the power of." So `(1 + 0.07)^10` means "1.07 multiplied by itself 10 times."

::: {.panel-tabset}

## R

```{r}
#| label: compound-r
future_value <- function(present_value, rate, years) {
  present_value * (1 + rate)^years
}

rule_of_72 <- function(annual_return_pct) {
  72 / annual_return_pct
}

# scalar: $10,000 at 7% for 10 years
future_value(present_value = 10000, rate = 0.07, years = 10)

# how long until your money doubles at 7%?
rule_of_72(annual_return_pct = 7)

# vectorized: same $10,000 at 7% across multiple horizons
tibble(years = c(5, 10, 20, 30)) |>
  mutate(value = future_value(present_value = 10000, rate = 0.07, years = years))
```

## Python

```{python}
#| label: compound-py
def future_value(present_value, rate, years):
    return present_value * (1 + rate) ** years

def rule_of_72(annual_return_pct):
    return 72 / annual_return_pct

# scalar: $10,000 at 7% for 10 years
print(future_value(present_value=10000, rate=0.07, years=10))

# how long until your money doubles at 7%?
print(rule_of_72(annual_return_pct=7))

# vectorized: numpy lets us pass an array of horizons in one call
horizons = np.array([5, 10, 20, 30])
future_value(present_value=10000, rate=0.07, years=horizons)
```

:::

### Opportunity Cost Math {.unnumbered}

This is Felix's signature framework. **Every dollar spent isn't just that dollar; it's what that dollar could have become.**

**Formula:** Cost × (1 + r)^n = Opportunity cost in future dollars

Example: A $40,000 new car vs. a $15,000 used car. The $25,000 difference, invested at 7% real return over 30 years:

$25,000 × (1.07)^30 = **~$190,000**

That's not a car payment difference; that's a retirement chapter.

::: {.panel-tabset}

## R

```{r}
#| label: opp-cost-r
# opportunity cost is just future value applied to the dollars you didn't invest
opportunity_cost <- function(amount, rate, years) {
  future_value(present_value = amount, rate = rate, years = years)
}

# $25,000 gap between two car choices, invested at 7% for 30 years
opportunity_cost(amount = 25000, rate = 0.07, years = 30)
```

## Python

```{python}
#| label: opp-cost-py
# opportunity cost is just future value applied to the dollars you didn't invest
def opportunity_cost(amount, rate, years):
    return future_value(present_value=amount, rate=rate, years=years)

# $25,000 gap between two car choices, invested at 7% for 30 years
opportunity_cost(amount=25000, rate=0.07, years=30)
```

:::

### Break-Even and Payback Calculations {.unnumbered}

Useful for evaluating subscriptions, refinances, and big purchases.

**Formula:** Upfront cost ÷ monthly savings = months to break even

Example: A $500 annual membership saves you $60/month → 500 ÷ 60 = **8.3 months to break even.** Worth it if you'll use it longer than that.

::: {.panel-tabset}

## R

```{r}
#| label: break-even-r
months_to_break_even <- function(upfront_cost, monthly_savings) {
  upfront_cost / monthly_savings
}

# a $500 membership that saves $60/month
months_to_break_even(upfront_cost = 500, monthly_savings = 60)
```

## Python

```{python}
#| label: break-even-py
def months_to_break_even(upfront_cost, monthly_savings):
    return upfront_cost / monthly_savings

# a $500 membership that saves $60/month
months_to_break_even(upfront_cost=500, monthly_savings=60)
```

:::

## Putting It All Together: A Budget That Actually Works

```{mermaid}
%%| fig-cap: 'Put It All Together'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Start[Build Your Budget] --> Step1[1. Track Spending<br/>for 30 Days]
    Step1 --> Step2[2. Categorize Fixed vs.<br/>Variable vs. Periodic]
    Step2 --> Step3[3. Calculate Current<br/>Savings Rate]
    Step3 --> Step4[4. Choose Framework<br/>Zero-Based or Percentage]
    Step4 --> Step5[5. Set Up Sinking Funds<br/>for Periodic Expenses]
    Step5 --> Step6[6. Automate Everything<br/>Investments First]
    Step6 --> Step7[7. Review Weekly,<br/>Adjust Monthly]
    Step7 --> Step8[8. Recalibrate Quarterly]
```

## The Bottom Line

A budget is just a **plan for your money in advance.** The best one:

1. **Matches your personality.** Don't fight your nature; work with it.
2. **Automates the boring stuff.** Investments, bills, sinking funds: set and forget.
3. **Leaves room for joy.** Sethi's guilt-free spending isn't optional; it's what makes budgets sustainable.
4. **Tracks one number above all: your savings rate.** Everything else is detail.
5. **Builds in margin for error.** Housel's principle: reality is messier than spreadsheets.

The math isn't hard. The categories aren't complicated. The tools are abundant. What's actually required is **the decision to be intentional**, and the patience to let small, consistent allocations compound into a life of financial freedom.

As Bogle reminded us about investing, the same applies to budgeting: **don't do something, just stand there.** Set up the system, then let it work.

