# Emergency Fund

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

An emergency fund is cash set aside to handle the unexpected without reaching for credit. It's the financial equivalent of a spare tire: you don't need it most of the time, but when you do, nothing else will do the job.

The standard target is **3 to 6 months of essential expenses**, parked somewhere safe and liquid. This chapter covers:

- How much you actually need
- Where to keep it (and where not to)
- How to build it from scratch
- How to rebuild it after a withdrawal

## Why It Matters

Most financial derailments don't start with bad investing. They start with a single unexpected expense — a car repair, a medical bill, a job loss — that gets charged to a credit card because there's no cash available.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Without an Emergency Fund'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Crisis([Unexpected Expense]) --> NoFund{Emergency Fund?}
    NoFund -->|No| CreditCard[Charge It to<br>Credit Card]
    NoFund -->|Yes| EFund[Draw from<br>Emergency Fund]
    CreditCard --> Interest[Pay 20-24% Interest<br>for Months/Years]
    CreditCard --> Stress[Financial Stress<br>& Anxiety]
    EFund --> Solved[Problem Solved<br>No Debt Added]
    EFund --> Rebuild[Rebuild the Fund<br>Over Next Few Months]
    style Crisis fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style NoFund fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style CreditCard fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Interest fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Stress fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style EFund fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Solved fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Rebuild fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
```

Housel's framing is useful here: an emergency fund isn't about the return on the money. It's about buying *options* — the ability to not be forced into a bad financial decision at a bad time. That optionality is worth more than any interest rate \cite{housel2020psychology}.

## How Much Do You Actually Need?

"3 to 6 months of expenses" is the standard answer, and it's a reasonable starting range. But it isn't a single number, because emergencies don't affect everyone equally.

### The Essential Expenses Baseline {.unnumbered}

Start by calculating your *essential expenses* — what it actually costs to keep the lights on. This is not your total spending; it's the floor.

| Category | Examples |
|---|---|
| **Housing** | Rent or mortgage, renter's/homeowner's insurance |
| **Food** | Groceries (not dining out) |
| **Utilities** | Electricity, water, gas, internet |
| **Transportation** | Car payment, insurance, fuel, transit pass |
| **Health** | Insurance premiums, critical medications |
| **Minimum debt payments** | Credit card minimums, loan payments |

Entertainment, subscriptions, gym memberships, and dining out are *not* essential expenses for this calculation. In a true emergency, you can cut them. Your emergency fund is sized to cover the life you *need*, not the one you currently live.

### Adjusting for Your Situation {.unnumbered}

The right target depends on how much income risk you carry. More risk = more months.

```{mermaid}
%%| fig-cap: 'How Many Months Do You Need?'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Start([How many months?]) --> Job{Job Stability?}
    Job -->|Stable<br>single income| Dep{Dependents?}
    Job -->|Variable / Self-Employed| More[Lean toward<br>6–9 months]
    Dep -->|No| Insure{Good health &<br>disability coverage?}
    Dep -->|Yes: kids, aging parents| SixPlus[6 months minimum]
    Insure -->|Yes| ThreeToSix[3–4 months<br>is sufficient]
    Insure -->|No| SixMonths[Closer to<br>6 months]
    style Start fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Job fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style More fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Dep fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style SixPlus fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Insure fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style ThreeToSix fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style SixMonths fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
```

A two-income household where both partners work stable salaried jobs has a lower emergency risk than a self-employed single parent. The fund should reflect that difference.

```{r}
#| label: ef-size-rec
#| echo: false
#| results: asis
#| eval: true
co_box(color = "g",
  hsize = "1.15",
  fold = FALSE,
  size = "1.10",
  header = "If you're just starting out",
  contents = "Don't let the 3-to-6 month target paralyze you. A $1,000 starter fund is meaningfully better than zero — it handles most routine emergencies (car repair, small medical bills, appliance replacement) and gives you a buffer while you pay down high-interest debt. Once toxic debt is cleared, build toward your full target.")
```

### The Starter Fund First {.unnumbered}

Sethi's approach is practical: build a $1,000 starter emergency fund immediately, even while paying down debt. Once high-interest debt is gone, shift the same monthly dollars you were paying toward debt into filling the emergency fund to its full target.

The sequence matters:

1. Build $1,000 starter fund
2. Eliminate credit card and high-interest debt (see [Managing Debt](managing_debt.qmd))
3. Build full 3–6 month fund

## Where to Keep It (and Where Not To)

An emergency fund has two jobs: stay safe and be available. This rules out most places you might initially consider.

```{mermaid}
%%| fig-cap: 'Where to Park Your Emergency Fund'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart LR
    Cash([Your Emergency<br>Fund]) --> Good["GOOD OPTIONS"]
    Cash --> Bad["BAD OPTIONS"]

    Good --> HYSA["High-Yield Savings Account<br>(HYSA)"]
    Good --> MMF["Money Market Fund<br>(at a brokerage)"]
    Good --> CheckFed["Federally Insured<br>Credit Union"]

    Bad --> Checking["Regular Checking Account<br>(too easy to spend)"]
    Bad --> Market["Stock Market<br>(too volatile)"]
    Bad --> CD["Long-Term CDs<br>(penalties for early withdrawal)"]
    Bad --> Cash2["Physical Cash at Home<br>(no interest, risk of loss)"]

    style Cash fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Good fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Bad fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style HYSA fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style MMF fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style CheckFed fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Checking fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Market fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style CD fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Cash2 fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
```

### The Right Account {.unnumbered}

A **high-yield savings account (HYSA)** at an FDIC-insured bank is the default recommendation. It earns meaningfully more than a standard checking account while keeping the money one business day away. Credit union equivalents covered by NCUA insurance work the same way.

See [High-Yield Accounts](high_yield_accounts.qmd) for a full breakdown of how to find and open one.

### Why Not the Stock Market? {.unnumbered}

This is the most common mistake people make with emergency funds. The logic seems sound: "*my money would earn more in index funds.*" The problem is sequencing risk.

If you lose your job in February and the stock market is down 25% at the same moment — which historically happens; recessions cause both — you're forced to sell at the worst possible time. The emergency fund's job isn't to grow wealth; it's to be there when you need it, at full value, on demand.

[Ben Felix approaches this through the lens of human capital.]{style="color:#48a56a; font-weight: bold;"} For most, your ability to earn an income is your largest asset. An emergency fund is a hedge against the volatility of that human capital. If your income is highly variable (e.g., sales commissions, freelancing, or business ownership), you should lean toward a larger cash buffer to protect your ability to maintain your lifestyle during lean periods.

```{r}
#| label: ef-invest-warning
#| echo: false
#| results: asis
#| eval: true
co_box(color = "o",
  hsize = "1.15",
  fold = FALSE,
  size = "1.10",
  header = "The temptation to invest it",
  contents = "Every bull market produces the same question: 'Why am I sitting on cash earning 4% when the market is up 20%?' The answer is that emergency funds serve a different function than investments. Conflating them exposes you to the exact sequence-of-returns risk you're trying to insure against. Keep your emergency fund in cash. Period.")
```

## Building Your Emergency Fund

For most people, building the fund from zero to 3–6 months takes 12–24 months of consistent contribution. The chart below shows the monthly balance for three different monthly contribution amounts toward a $15,000 target.

```{r}
#| label: ef-build-bar-model
#| echo: false
#| message: false 
#| warning: false
#| fig-cap: "Months to reach a $15,000 emergency fund at three contribution levels"
#| fig-asp: 0.618
library(ggplot2)
library(dplyr)

target <- 15000
contribs <- c(300, 500, 750)
colors <- c("#86ddcd", "#0e9aa7", "#48a56a")

df_build <- purrr::map2_dfr(contribs, colors, function(c, col) {
  months <- ceiling(target / c)
  data.frame(
    monthly   = paste0("$", c, "/mo"),
    months    = months,
    fill      = col
  )
})

df_build$monthly <- factor(df_build$monthly, levels = df_build$monthly)

ggplot(df_build, aes(x = monthly, y = months, fill = monthly)) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = paste0(months, " months")),
    vjust = -0.4, fontface = "bold", size = 4
  ) +
  scale_fill_manual(values = setNames(df_build$fill, df_build$monthly)) +
  scale_y_continuous(limits = c(0, 55), labels = function(x) paste0(x, " mo")) +
  labs(x = "Monthly contribution", y = NULL,
       subtitle = "Target: $15,000 (approx. 3 months of $5,000/month expenses)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

The pattern is simple: doubling the monthly contribution roughly halves the time to goal.

### Automate It {.unnumbered}

The most reliable way to build the fund is to treat it exactly like a bill. Set up an automatic transfer to your HYSA on payday, before you have a chance to spend the money. See [Automating Savings](automating_savings.qmd) for the full setup playbook.

### Where to Find the Money {.unnumbered}

If you're reading this and thinking "*I don't have anything extra to save,*" the answer is usually a combination of small cuts applied consistently:

- Pause one subscription
- Cook at home one extra night per week
- Redirect a tax refund directly to the fund

None of these feel dramatic. That's the point. The fund gets built one unexciting month at a time.

## How to Rebuild After a Withdrawal

Using the emergency fund isn't a failure — it's the fund doing its job. The mistake is treating a withdrawal as a one-time event rather than a signal to rebuild immediately.

```{mermaid}
%%| fig-cap: 'Rebuild Protocol'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Use([Emergency: Fund Used]) --> Assess[Assess: How much<br>was withdrawn?]
    Assess --> Resume[Resume automatic<br>contributions immediately]
    Resume --> Temp{Can you temporarily<br>increase contributions?}
    Temp -->|Yes| Boost[Increase monthly<br>contribution until rebuilt]
    Temp -->|No| Steady[Maintain existing<br>contribution; stay patient]
    Boost --> Target([Fund Restored])
    Steady --> Target
    style Use fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Assess fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Resume fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Temp fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Boost fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Steady fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Target fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
```

Two rules for the rebuild phase:

1. **Don't stop other savings to rebuild faster.** The automatic contributions will get you there. Don't interrupt retirement account contributions, especially if you have an employer match.

2. **Don't restart debt payoff conversations during the rebuild.** Your priority order right now is: minimum payments on everything, then rebuild the emergency fund, then return to accelerated debt payoff.

## Math for Emergency Funds {#math-for-emergency-funds}

The calculations here are simple, but writing them down forces you to confront the actual numbers.

### Your Target Amount {.unnumbered}

Your emergency fund target is just essential monthly expenses multiplied by the number of months you want to cover.

The chart below shows how the target grows as you add months — each bar is the previous one plus one more month of expenses.

```{r}
#| label: ef-target-bar-model
#| echo: false
#| fig-cap: "Emergency fund target at $3,500/month essential expenses"
#| fig-asp: 0.50
library(ggplot2)
monthly_exp <- 3500
months_vec  <- 1:6
df_target   <- data.frame(
  months = factor(paste0(months_vec, " mo"), levels = paste0(months_vec, " mo")),
  target = monthly_exp * months_vec,
  fill   = c("#86ddcd", "#86ddcd", "#0e9aa7", "#0e9aa7", "#48a56a", "#48a56a")
)
ggplot(df_target, aes(x = months, y = target, fill = months)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0("$", format(target, big.mark = ","))),
    vjust = -0.4, fontface = "bold", size = 3.8
  ) +
  scale_fill_manual(values = setNames(df_target$fill, df_target$months)) +
  scale_y_continuous(
    limits = c(0, monthly_exp * 7.5),
    labels = function(x) paste0("$", format(x, big.mark = ","))
  ) +
  labs(x = "Months of expenses covered", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

**Formula:** Target = Monthly Essential Expenses × Months of Coverage

Example: \$3,500/month in essential expenses × 4 months = [\$14,000 target.]{style="color:#48a56a; font-weight: bold;"}

::: panel-tabset
## R

```{r}
#| label: ef-target-r
emergency_fund_target <- function(monthly_expenses, months = 3) {
  monthly_expenses * months
}

# $3,500/month in essential expenses, 3-month and 6-month targets
emergency_fund_target(monthly_expenses = 3500, months = 3)
emergency_fund_target(monthly_expenses = 3500, months = 6)
```

## Python

```{python}
#| label: ef-target-py
def emergency_fund_target(monthly_expenses, months=3):
    return monthly_expenses * months

# $3,500/month in essential expenses, 3-month and 6-month targets
print(emergency_fund_target(monthly_expenses=3500, months=3))
print(emergency_fund_target(monthly_expenses=3500, months=6))
```
:::

### Months to Reach Your Target {.unnumbered}

Given your current balance and a monthly contribution, how long until you reach the target?

**Formula:** Months to Goal = (Target − Current Balance) ÷ Monthly Contribution

Example: \$14,000 target, \$2,000 already saved, contributing \$500/month → (14,000 − 2,000) ÷ 500 = [24 months.]{style="color:#48a56a; font-weight: bold;"}

::: panel-tabset
## R

```{r}
#| label: ef-months-r
months_to_ef_goal <- function(target, current_balance, monthly_contribution) {
  remaining <- max(0, target - current_balance)
  ceiling(remaining / monthly_contribution)
}

# $14,000 target, $2,000 saved, $500/month contribution
months_to_ef_goal(target = 14000, current_balance = 2000, monthly_contribution = 500)
```

## Python

```{python}
#| label: ef-months-py
import math

def months_to_ef_goal(target, current_balance, monthly_contribution):
    remaining = max(0, target - current_balance)
    return math.ceil(remaining / monthly_contribution)

# $14,000 target, $2,000 saved, $500/month contribution
print(months_to_ef_goal(target=14000, current_balance=2000, monthly_contribution=500))
```
:::

### Contribution Needed by a Deadline {.unnumbered}

Working backwards: if you want to reach your target in a specific number of months, how much do you need to save each month?

**Formula:** Monthly Contribution = (Target − Current Balance) ÷ Months Remaining

Example: You want a full \$14,000 fund in 18 months and have \$2,000 now → (14,000 − 2,000) ÷ 18 = [\$667/month.]{style="color:#48a56a; font-weight: bold;"}

::: panel-tabset
## R

```{r}
#| label: ef-contrib-r
monthly_contribution_needed <- function(target, current_balance, months) {
  remaining <- max(0, target - current_balance)
  ceiling(remaining / months)
}

# $14,000 target, $2,000 saved, want to finish in 18 months
monthly_contribution_needed(target = 14000, current_balance = 2000, months = 18)
```

## Python

```{python}
#| label: ef-contrib-py
def monthly_contribution_needed(target, current_balance, months):
    remaining = max(0, target - current_balance)
    return math.ceil(remaining / months)

# $14,000 target, $2,000 saved, want to finish in 18 months
print(monthly_contribution_needed(target=14000, current_balance=2000, months=18))
```
:::

### What the Fund Earns While You Build It {.unnumbered}

Keeping the fund in a HYSA means it earns interest as it grows. This won't dramatically change your timeline, but it's real money.

**Formula:** Future Value = Balance × (1 + monthly rate)\^months

Where monthly rate = annual rate ÷ 12.

Example: \$14,000 in a HYSA at 4.5% APY, untouched for 12 months → [\~\$14,630.]{style="color:#48a56a; font-weight: bold;"} The interest pays for a car repair without reducing your fund.

::: panel-tabset
## R

```{r}
#| label: ef-growth-r
ef_growth <- function(balance, annual_rate, months) {
  monthly_rate <- annual_rate / 12
  balance * (1 + monthly_rate)^months
}

# $14,000 at 4.5% APY for 12 months
ef_growth(balance = 14000, annual_rate = 0.045, months = 12)
```

## Python

```{python}
#| label: ef-growth-py
def ef_growth(balance, annual_rate, months):
    monthly_rate = annual_rate / 12
    return balance * (1 + monthly_rate) ** months

# $14,000 at 4.5% APY for 12 months
print(ef_growth(balance=14000, annual_rate=0.045, months=12))
```
:::

## Putting It All Together

```{mermaid}
%%| fig-cap: 'Build Your Emergency Fund'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Start(["Build Your<br>Emergency Fund"]) --> Step1("Step 1. Calculate monthly<br>essential expenses")
    Step1 --> Step2("Step 2. Choose your target:<br>3–6 months, based on risk")
    Step2 --> Step3("Step 3. Open a<br>dedicated HYSA")
    Step3 --> Step4("Step 4. Set up automatic<br>monthly transfer on payday")
    Step4 --> Step5("Step 5. Build $1,000 starter<br>fund first if in debt")
    Step5 --> Step6("Step 6. Pay off high-interest<br>debt, then fill full fund")
    Step6 --> Step7("Step 7. Once funded, leave it<br>alone — it's not savings")
    Step7 --> Step8(["Step 8. Rebuild immediately<br>after any withdrawal"])
    style Start fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step1 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step2 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step3 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step4 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step5 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step6 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step7 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Step8 fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
```

## Key takeaways

An emergency fund is not an investment. It's insurance — and the cost of that insurance is the interest you *don't* earn by keeping it in cash instead of equities. That's a reasonable price for what it buys.

1.  [Size it to your actual risk.]{style="color:#48a56a; font-weight: bold;"} Stable salaried job with good coverage? Three months is fine. Variable income or dependents? Six months or more.
2.  [Keep it separate.]{style="color:#48a56a; font-weight: bold;"} A dedicated HYSA with a slight friction to transfer prevents casual spending from eroding the fund.
3.  [Build it in stages.]{style="color:#48a56a; font-weight: bold;"} $1,000 starter fund first, then full fund after high-interest debt is cleared.
4.  [Automate the contribution.]{style="color:#48a56a; font-weight: bold;"} Set it on payday and treat it like a bill until the fund is full.
5.  [Rebuild immediately after using it.]{style="color:#48a56a; font-weight: bold;"} The fund only works if it's there the *next* time you need it.

As Housel writes, the point of financial margin isn't to optimize — it's to survive. An emergency fund doesn't make you rich. It keeps a bad month from becoming a bad decade.
