# Welcome! {.unnumbered}

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

Welcome to the **Personal Finance field manual**, a working reference for the day-to-day decisions that shape your financial life: budgeting, savings, borrowing, investing, taxes, insurance, and planning.


```{r}
#| label: intro-disclaimer
#| echo: false
#| results: asis
#| eval: true
co_box(color = "r", 
  header = "NOTE", 
  contents = "This primary audience for this book is my children. It contains some of the basic principles I was taught (and others I've learned painfully) about managing personal finances.")
```


This book aims to answer questions like:

  - "*Where is my money going each month?*"

  - "*How much should I have in an emergency fund?*"

  - "*Should I pay down debt or invest first?*"

  - "*Which retirement account makes sense for me?*"

  - "*How do I pick an index fund?*"

## How this book is organized

The book is divided into five parts:

1. [**Foundations**](foundations.qmd) -- budgeting, tracking spending, emergency funds, and debt.
2. [**Savings**](savings.qmd) -- savings strategies, high-yield accounts, and automation.
3. [**Investing**](investing.qmd)  -- the basics, retirement accounts, index funds, and asset allocation.
4. [**Taxes & insurance**](taxes_insurance.qmd) -- the protective layer around your finances.
5. [**Planning**](planning.qmd) -- goals, major purchases, and estate planning.

The five sections of the book map onto the **four levers**: 

1. *Foundations* sets up income and spending  
2. *Savings* turns the gap between them into a buffer  
3. *Investing* puts that buffer to work   
4. *Taxes & insurance* and *Planning* sit around the outside  

Each chapter is short and self-contained so you can jump in where you need to.

## Code

If you'd like to follow along or perform any of the math yourself, the examples use both **R** (with the `tidyverse`) and **Python** (with `pandas` / `NumPy`). Pick whichever you prefer -- the underlying ideas are the same.

```{r}
#| label: ex-r
1000 * (1 + 0.05)^10  # future value of $1,000 at 5% for 10 years
```

```{python}
#| label: ex-py
1000 * (1 + 0.05) ** 10
```

## About the author

Martin Frigaard writes about R, data, and -- now -- personal finance.
