# Welcome! {.unnumbered}

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

Welcome to my **Personal Finance field manual**, a working reference for the day-to-day decisions that shape your financial life: budgeting, savings, borrowing, investing, taxes, insurance, and planning.


This book aims to answer questions like:

  - "*Where is my money going each month?*"

  - "*How much should I have in an emergency fund?*"

  - "*Should I pay down debt or invest first?*"

  - "*Which retirement account makes sense for me?*"

  - "*How do I pick an index fund?*"

```{r}
#| label: llm-disclaiamer
#| echo: false 
#| results: asis
#| eval: true
co_box(
  color = "r",
  header = "Who wrote this?",
  contents = "Given the current landscape of online resources and the impressive capabilities of AI/large language models (LLMs), I feel it’s important to clarify the authorship of this book. It seems strange to have to include this in anything I write, but it’s something that, if I were reading a book, I’d like to know. 
  
**I wrote all of the narrative, commentary, and explanations.** I used LLMs (mostly Claude) to research specific topics (like FDIC/NCUA coverage), but the thoughts in the book are my own. Hopefully, the typos and/or grammatical errors serve as a signature of authenticity. 

**The code in this book was created with the assistance of Claude, and this applies only to the specific chapters explicitly noted.** Generally, I’d write a ‘first draft’ of a code example, then ask Claude how to improve it. Sometimes I agreed with its suggestions (and included them), other times I revised the code myself."
)
```

## How this book is organized

The book is divided into five parts:

1. [**Foundations**](foundations.qmd): budgeting, tracking spending, emergency funds, and debt.
2. [**Savings**](savings.qmd): savings strategies, high-yield accounts, and automation.
3. [**Investing**](investing.qmd): the basics, retirement accounts, index funds, and asset allocation.
4. [**Taxes & insurance**](taxes_insurance.qmd): the protective layer around your finances.
5. [**Planning**](planning.qmd): goals, major purchases, and estate planning.

The five sections of the book map onto the **four levers**: 

1. *Foundations* sets up income and spending  
2. *Savings* turns the gap between them into a buffer  
3. *Investing* puts that buffer to work   
4. *Taxes & insurance* and *Planning* sit around the outside  

Each chapter is short and self-contained so you can jump in where you need to.

## Math

Whatever math is covered in this book can be done with a calculator, spreadsheet, or some basic programming skills. If you're interested in following along or performing any of the calculations yourself, the examples use both **R** and **Python**. 

Pick whichever you prefer; the underlying ideas are the same.

```{r}
#| label: ex-r
1000 * (1 + 0.05)^10  # future value of $1,000 at 5% for 10 years
```

```{python}
#| label: ex-py
1000 * (1 + 0.05) ** 10
```

```{r}
#| label: intro-disclaimer
#| echo: false
#| results: asis
#| eval: true
co_box(color = "g", 
  header = "NOTE", 
  contents = "This primary audience for this book is my children. It contains some of the basic principles I was taught (and others I've learned painfully) about managing personal finances.")
```

## About the author

Martin Frigaard writes about R packages, Shiny app development, and, now, personal finance.
