# Investing Basics

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

A short tour of the concepts behind every investment decision.

## Compounding

Compounding is interest earning interest. Over long horizons, it does the
heavy lifting.

```{r}
#| label: compounding-r
years   <- 0:40
balance <- 10000 * (1.07)^years
plot(years, balance, type = "l",
     xlab = "Years", ylab = "Balance ($)",
     main = "$10,000 at 7% compounded annually")
```

```{python}
#| label: compounding-py
years = list(range(0, 41))
balance = [10000 * (1.07) ** y for y in years]
balance[-1]  # ending balance after 40 years
```

**Formula:** FV = PV × (1 + r)\^n

Where: FV = future value, PV = present value, r = annual return, n = years

**The shortcut you should memorize is the Rule of 72:**

> Years to double your money = 72 ÷ annual return %

At 7% return, money doubles every \~10 years. At 10%, every \~7 years. This single insight reframes every spending decision.

*See [Math for Investing Basics](#math-for-investing-basics) below for the `future_value()` and `rule_of_72()` functions in R and Python, plus the math for regular contributions and inflation-adjusted returns.*

## Risk vs. return

Higher expected return = higher expected volatility. Choose a level you can
[stick with]{style="color:#48a56a; font-weight: bold;"} through downturns, not just one that looks attractive on a
spreadsheet.

## Diversification

Don't bet on one company, one sector, or one country. Broad index funds
make this nearly free.

## Math for Investing Basics {#math-for-investing-basics}

Investing math is mostly one idea, compounding, applied in a few different ways. Each calculation below is written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter.

In code, the `^` symbol (R) and `**` symbol (Python) both mean "raise to the power of." So `(1 + 0.07)^10` means "1.07 multiplied by itself 10 times."

### Future Value and the Rule of 72 {.unnumbered}

Each bar below is the previous one multiplied by the same growth factor. The result is not a straight line — it curves upward because each year's growth is applied to a larger base.

```{r}
#| label: compound-bar-model
#| echo: false
#| fig-cap: "$10,000 growing at 7% — value at each checkpoint"
#| fig-asp: 0.50
library(ggplot2)
yrs  <- c(0, 5, 10, 20, 30)
vals <- round(10000 * (1.07)^yrs)
bar_colors <- c("#86ddcd", "#5ec8be", "#0e9aa7", "#e07070", "#f44242")
df_fv <- data.frame(
  year  = factor(paste0("Year ", yrs), levels = paste0("Year ", yrs)),
  value = vals,
  fill  = bar_colors
)
ggplot(df_fv, aes(x = year, y = value, fill = fill)) +
  geom_col(width = 0.6, colour = "#1C1C1E", linewidth = 0.4) +
  geom_text(
    aes(label = paste0("$", formatC(value, format = "d", big.mark = ","))),
    vjust = -0.4, fontface = "bold", size = 3.5
  ) +
  scale_fill_identity() +
  scale_y_continuous(
    limits = c(0, max(vals) * 1.18),
    labels = function(x) paste0("$", round(x / 1000), "k")
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

Multiplying by (1 + r) once gives next year's value; multiplying by it n times gives the value after n years — that's the formula.

**Formula:** FV = PV × (1 + r)\^n

Where: FV = future value, PV = present value (today's amount), *r* = annual return, *n* = years.

The companion shortcut is the **Rule of 72**: divide 72 by the return percentage to estimate how many years it takes your money to double.

> Years to double ≈ 72 ÷ annual return %

Example: \$10,000 at 7% for 10 years grows to \~\$19,672, and at 7% money doubles roughly every 10 years.

::: panel-tabset
## R

```{r}
#| label: compound-r
future_value <- function(present_value, rate, years) {
  present_value * (1 + rate)^years
}

rule_of_72 <- function(annual_return_pct) {
  72 / annual_return_pct
}

# scalar: $10,000 at 7% for 10 years
future_value(present_value = 10000, rate = 0.07, years = 10)

# how long until your money doubles at 7%?
rule_of_72(annual_return_pct = 7)

# vectorized: same $10,000 at 7% across multiple horizons
data.frame(
  years = c(5, 10, 20, 30),
  value = future_value(present_value = 10000, rate = 0.07, years = c(5, 10, 20, 30))
)
```

## Python

```{python}
#| label: compound-py
import numpy as np

def future_value(present_value, rate, years):
    return present_value * (1 + rate) ** years

def rule_of_72(annual_return_pct):
    return 72 / annual_return_pct

# scalar: $10,000 at 7% for 10 years
print(future_value(present_value=10000, rate=0.07, years=10))

# how long until your money doubles at 7%?
print(rule_of_72(annual_return_pct=7))

# vectorized: numpy lets us pass an array of horizons in one call
horizons = np.array([5, 10, 20, 30])
future_value(present_value=10000, rate=0.07, years=horizons)
```
:::

### Future Value of Regular Contributions {.unnumbered}

Investing a fixed amount every month builds wealth in two ways: the money you put in, and the growth earned on all previous contributions. The chart below separates these two layers so you can see when growth overtakes contributions.

```{r}
#| label: fv-series-bar-model
#| echo: false
#| fig-cap: "$500/month contributions at 7% — contributed vs. gains"
#| fig-asp: 0.50
library(ggplot2)
yrs_vec    <- c(10, 20, 30)
annual_pmt <- 500 * 12
contributed <- annual_pmt * yrs_vec
total       <- annual_pmt * ((1.07^yrs_vec - 1) / 0.07)
gains       <- total - contributed
df_series <- data.frame(
  year        = factor(paste0("Year ", yrs_vec),
                       levels = paste0("Year ", yrs_vec)),
  contributed = contributed,
  gains       = gains
)
df_long <- data.frame(
  year    = rep(df_series$year, 2),
  amount  = c(df_series$contributed, df_series$gains),
  layer   = factor(rep(c("Money contributed", "Investment gains"),
                       each = nrow(df_series)),
                   levels = c("Investment gains", "Money contributed"))
)
totals_lab <- data.frame(
  year  = df_series$year,
  total = contributed + gains
)
ggplot(df_long, aes(x = year, y = amount, fill = layer)) +
  geom_col(width = 0.6, colour = "#1C1C1E", linewidth = 0.4) +
  geom_text(
    data = totals_lab,
    aes(x = year, y = total,
        label = paste0("$", formatC(round(total), format = "d", big.mark = ","))),
    inherit.aes = FALSE,
    vjust = -0.4, fontface = "bold", size = 3.5
  ) +
  scale_fill_manual(
    values = c("Money contributed" = "#0e9aa7", "Investment gains" = "#86ddcd")
  ) +
  scale_y_continuous(
    limits = c(0, max(contributed + gains) * 1.15),
    labels = function(x) paste0("$", round(x / 1000), "k")
  ) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "top"
  )
```

The formula sums all those contribution bars growing at different rates — the annuity formula collapses that sum into a single calculation.

**Formula:** FV = PMT × [((1 + r)\^n − 1) ÷ r]

Where: PMT = the amount contributed each period, *r* = the return per period, *n* = the number of periods.

Example: \$500/month for 30 years (360 months) at a 7% annual return (\~0.583% per month) grows to \~\$610,000, of which only \$180,000 is money you actually put in.

::: panel-tabset
## R

```{r}
#| label: fv-series-r
future_value_series <- function(contribution, rate, periods) {
  contribution * (((1 + rate)^periods - 1) / rate)
}

# $500/month for 30 years at a 7% annual return (monthly rate = 0.07 / 12)
future_value_series(contribution = 500, rate = 0.07 / 12, periods = 30 * 12)
```

## Python

```{python}
#| label: fv-series-py
def future_value_series(contribution, rate, periods):
    return contribution * (((1 + rate) ** periods - 1) / rate)

# $500/month for 30 years at a 7% annual return (monthly rate = 0.07 / 12)
future_value_series(contribution=500, rate=0.07 / 12, periods=30 * 12)
```
:::

### Real (Inflation-Adjusted) Return {.unnumbered}

A 7% nominal return sounds good, but inflation is quietly shrinking the purchasing power of every dollar. The chart below shows the three quantities: the stated return, the inflation cost, and what actually remains.

```{r}
#| label: real-return-bar-model
#| echo: false
#| fig-cap: "Nominal return, inflation, and real return"
#| fig-asp: 0.50
library(ggplot2)
df_rr <- data.frame(
  label  = factor(
    c("Nominal return (7%)", "Inflation (3%)", "Real return (~3.9%)"),
    levels = c("Nominal return (7%)", "Inflation (3%)", "Real return (~3.9%)")
  ),
  value  = c(7, 3, (1.07 / 1.03 - 1) * 100),
  fill   = c("#0e9aa7", "#f44242", "#48a56a")
)
ggplot(df_rr, aes(x = label, y = value, fill = fill)) +
  geom_col(width = 0.55, colour = "#1C1C1E", linewidth = 0.4) +
  geom_text(
    aes(label = paste0(round(value, 1), "%")),
    vjust = -0.4, fontface = "bold", size = 4
  ) +
  scale_fill_identity() +
  scale_y_continuous(
    limits = c(0, 9),
    labels = function(x) paste0(x, "%")
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

The real return is what's left after inflation takes its slice — not a simple subtraction, but a ratio, because the percentages compound.

**Formula:** Real Return = (1 + nominal) ÷ (1 + inflation) − 1

Example: (1.07 ÷ 1.03) − 1 = \~3.88%, slightly less than the 4% you'd get by simple subtraction.

::: panel-tabset
## R

```{r}
#| label: real-return-r
real_return <- function(nominal, inflation) {
  (1 + nominal) / (1 + inflation) - 1
}

# 7% nominal return against 3% inflation
real_return(nominal = 0.07, inflation = 0.03)
```

## Python

```{python}
#| label: real-return-py
def real_return(nominal, inflation):
    return (1 + nominal) / (1 + inflation) - 1

# 7% nominal return against 3% inflation
real_return(nominal=0.07, inflation=0.03)
```
:::
