# Investing Basics

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

A short tour of the concepts behind every investment decision.

## Compounding

Compounding is interest earning interest. Over long horizons, it does the
heavy lifting.

```{r}
#| label: compounding-r
years   <- 0:40
balance <- 10000 * (1.07)^years
plot(years, balance, type = "l",
     xlab = "Years", ylab = "Balance ($)",
     main = "$10,000 at 7% compounded annually")
```

```{python}
#| label: compounding-py
years = list(range(0, 41))
balance = [10000 * (1.07) ** y for y in years]
balance[-1]  # ending balance after 40 years
```

**Formula:** FV = PV × (1 + r)\^n

Where: FV = future value, PV = present value, r = annual return, n = years

**The shortcut you should memorize is the Rule of 72:**

> Years to double your money = 72 ÷ annual return %

At 7% return, money doubles every \~10 years. At 10%, every \~7 years. This single insight reframes every spending decision.

In code, the `^` symbol (R) and `**` symbol (Python) both mean "raise to the power of." So `(1 + 0.07)^10` means "1.07 multiplied by itself 10 times."

::: panel-tabset
## R

```{r}
#| label: compound-r
future_value <- function(present_value, rate, years) {
  present_value * (1 + rate)^years
}

rule_of_72 <- function(annual_return_pct) {
  72 / annual_return_pct
}

# scalar: $10,000 at 7% for 10 years
future_value(present_value = 10000, rate = 0.07, years = 10)

# how long until your money doubles at 7%?
rule_of_72(annual_return_pct = 7)

# vectorized: same $10,000 at 7% across multiple horizons
data.frame(
  years = c(5, 10, 20, 30),
  value = future_value(present_value = 10000, rate = 0.07, years = c(5, 10, 20, 30))
)
```

## Python

```{python}
#| label: compound-py
import numpy as np

def future_value(present_value, rate, years):
    return present_value * (1 + rate) ** years

def rule_of_72(annual_return_pct):
    return 72 / annual_return_pct

# scalar: $10,000 at 7% for 10 years
print(future_value(present_value=10000, rate=0.07, years=10))

# how long until your money doubles at 7%?
print(rule_of_72(annual_return_pct=7))

# vectorized: numpy lets us pass an array of horizons in one call
horizons = np.array([5, 10, 20, 30])
future_value(present_value=10000, rate=0.07, years=horizons)
```
:::

## Risk vs. return

Higher expected return = higher expected volatility. Choose a level you can
**stick with** through downturns, not just one that looks attractive on a
spreadsheet.

## Diversification

Don't bet on one company, one sector, or one country. Broad index funds
make this nearly free.
