# Major Purchases

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>


Houses, cars, and weddings are where most people accidentally undo years
of saving. The mechanics aren't complicated; the discipline is.

  - Down payment math
  - Total cost of ownership (the sticker price is the smallest piece)
  - Financing vs. paying cash
  - Avoiding lifestyle creep after a raise

## Owning vs. Renting: The Most Debated Question

This is where [Ben Felix](https://rationalreminder.ca/)'s research is genuinely groundbreaking. The conventional wisdom ("renting is throwing money away") is one of the most persistent myths in personal finance.

### The Real Math {.unnumbered}

Felix has produced extensive analysis showing that **renting and investing the difference often matches or beats homeownership**, once you account for the full cost of owning.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Home Ownership'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    HomeCosts[True Cost<br>of Ownership] --> Mortgage[Mortgage<br>Interest]
    HomeCosts --> PropTax[Property Taxes<br>1-2% annually]
    HomeCosts --> Maintenance[Maintenance<br>~1% annually]
    HomeCosts --> Insurance[Insurance]
    HomeCosts --> Closing[Closing<br>Costs 3-5%]
    HomeCosts --> Opportunity2[Down Payment<br>Opportunity Cost]

    Mortgage --> UnrecovCost[Unrecoverable Costs<br>~5% of home<br>value per year]
    PropTax --> UnrecovCost
    Maintenance --> UnrecovCost
    Insurance --> UnrecovCost
    style HomeCosts fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Mortgage fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style PropTax fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Maintenance fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Insurance fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Closing fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Opportunity2 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style UnrecovCost fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
```

### Felix's "5% Rule" {.unnumbered}

A simplified version of Felix's framework: **the unrecoverable costs of owning are roughly 5% of the home's value per year** (property tax ~1%, maintenance ~1%, cost of capital ~3%).

**Rule of thumb:** Multiply the home's value by 5% and divide by 12. If you can rent an equivalent property for less than that, **renting and investing the difference is likely the better financial choice.**

Example: $500,000 home × 5% = $25,000/year = ~$2,080/month. If you can rent the same place for less, renting wins financially.

### When Buying Makes Sense {.unnumbered}

Buying isn't bad, but it should be a **lifestyle decision** more than a financial one. Buy when:

- You're confident you'll stay 7+ years (closing costs need time to amortize)
- You have a stable income and adequate emergency fund
- You value the autonomy, stability, and control
- The local rent-to-price ratio favors buying
- You won't be "house poor" (housing costs should stay under 28-30% of gross income)

### When Renting Makes Sense {.unnumbered}

- You may move within 5-7 years
- Your career requires geographic flexibility
- Local home prices are stretched relative to rents
- You'd rather invest the difference in a diversified portfolio
- You value flexibility, lower maintenance burden, and predictable costs

**The key insight from Felix:** renting is not "throwing money away" any more than buying groceries is throwing money away. **You're paying for shelter either way.** The question is which form of shelter delivers better financial and lifestyle outcomes for your situation.

### The Rent vs. Buy Calculation (Felix's 5% Rule) {.unnumbered}

This deserves its own line because it's the most consequential housing decision most people make.

**The 5% Rule (Ben Felix's framework):**

> Annual unrecoverable cost of owning ≈ Home Value × 5%

Broken down:

- Property tax: ~1%
- Maintenance: ~1%
- Cost of capital (opportunity cost of down payment + mortgage interest): ~3%

**Decision math:**

Monthly equivalent rent threshold = (Home Value × 0.05) ÷ 12

Example: $600,000 home → (600,000 × 0.05) ÷ 12 = **$2,500/month**

If you can rent the equivalent home for less than $2,500/month, renting is financially favorable (when you invest the difference). If rent exceeds that, buying tips ahead.

### Mortgage Amortization: The Front-Loaded Interest Reality {.unnumbered}

Mortgages are structured so you pay **mostly interest in early years.** Understanding this changes how you think about moving and refinancing.

**Key insight:** On a 30-year mortgage at 6%, roughly **75% of your first year's payments go to interest**, not principal.

**Rough breakdown of where payments go:**

| Years | Approx. % to Interest | Approx. % to Principal |
|-------|----------------------|------------------------|
| 1-5 | 70-80% | 20-30% |
| 6-15 | 50-70% | 30-50% |
| 16-25 | 25-50% | 50-75% |
| 26-30 | 5-25% | 75-95% |

This is why moving every 5 years is so financially destructive: you're essentially renting from the bank while paying ownership costs on top.

## Math for Major Purchases {#math-for-major-purchases}

Big purchases are where small percentage differences turn into tens of thousands of dollars. Each calculation below is written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter.

### The Rent-vs-Buy 5% Rule {.unnumbered}

Felix's rule estimates the annual *unrecoverable* cost of owning as about 5% of the home's value, then compares the monthly equivalent to your rent.

**Formula:** Monthly cost of owning ≈ (Home Value × 0.05) ÷ 12

Example: a \$600,000 home → (600,000 × 0.05) ÷ 12 = **\$2,500/month**. Rent below that, and renting-and-investing-the-difference tends to win.

::: panel-tabset
## R

```{r}
#| label: rent-buy-r
rent_buy_threshold <- function(home_value, unrecoverable_rate = 0.05) {
  (home_value * unrecoverable_rate) / 12
}

# $600,000 home at the standard 5% unrecoverable rate
rent_buy_threshold(home_value = 600000)
```

## Python

```{python}
#| label: rent-buy-py
def rent_buy_threshold(home_value, unrecoverable_rate=0.05):
    return (home_value * unrecoverable_rate) / 12

# $600,000 home at the standard 5% unrecoverable rate
rent_buy_threshold(home_value=600000)
```
:::

### The Monthly Mortgage Payment {.unnumbered}

This is the standard amortization formula behind every mortgage calculator.

**Formula:** M = P × [ r(1 + r)\^n ] ÷ [ (1 + r)\^n − 1 ]

Where: P = the loan principal, *r* = the monthly interest rate (annual ÷ 12), *n* = the total number of payments (years × 12).

Example: a \$480,000 loan (a \$600,000 home with 20% down) at 6% for 30 years → **\~\$2,878/month**.

::: panel-tabset
## R

```{r}
#| label: mortgage-payment-r
mortgage_payment <- function(principal, annual_rate, years) {
  r <- annual_rate / 12
  n <- years * 12
  principal * (r * (1 + r)^n) / ((1 + r)^n - 1)
}

# $480,000 loan at 6% over 30 years
mortgage_payment(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

```{python}
#| label: mortgage-payment-py
def mortgage_payment(principal, annual_rate, years):
    r = annual_rate / 12
    n = years * 12
    return principal * (r * (1 + r) ** n) / ((1 + r) ** n - 1)

# $480,000 loan at 6% over 30 years
mortgage_payment(principal=480000, annual_rate=0.06, years=30)
```
:::

### Where the First Year's Payments Go {.unnumbered}

The front-loaded-interest reality, in numbers: in month one your interest is simply the balance times the monthly rate, and only the rest reduces the principal.

**Formula:**

- First-month interest = Principal × (annual rate ÷ 12)
- First-month principal = Monthly Payment − First-month interest

Example: on that \$480,000 loan, month-one interest is \$2,400 of the \~\$2,878 payment, about **83% interest**, leaving only \~\$478 toward principal.

::: panel-tabset
## R

```{r}
#| label: first-payment-r
first_payment_split <- function(principal, annual_rate, years) {
  payment  <- mortgage_payment(principal, annual_rate, years)
  interest <- principal * (annual_rate / 12)
  data.frame(
    payment      = payment,
    interest     = interest,
    principal    = payment - interest,
    pct_interest = (interest / payment) * 100
  )
}

# first payment on a $480,000 loan at 6% over 30 years
first_payment_split(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

```{python}
#| label: first-payment-py
def first_payment_split(principal, annual_rate, years):
    payment = mortgage_payment(principal, annual_rate, years)
    interest = principal * (annual_rate / 12)
    return {
        "payment": payment,
        "interest": interest,
        "principal": payment - interest,
        "pct_interest": (interest / payment) * 100,
    }

# first payment on a $480,000 loan at 6% over 30 years
first_payment_split(principal=480000, annual_rate=0.06, years=30)
```
:::
