# Major Purchases

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>


A house is where most people accidentally undo years of saving. The mechanics aren't complicated; the discipline is.

  * Renting vs. buying
  * The true cost of owning (the sticker price is the smallest piece)
  * Monthly mortgage payments and how interest is front-loaded

## Owning vs. Renting: The Most Debated Question

This is where [Ben Felix](https://rationalreminder.ca/)'s research is genuinely groundbreaking. The conventional wisdom ("renting is throwing money away") is one of the most persistent myths in personal finance.

### The Real Math {.unnumbered}

Felix has produced extensive analysis showing that [renting and investing the difference often matches or beats homeownership]{style="color:#48a56a; font-weight: bold;"}, once you account for the full cost of owning.

```{mermaid}
%%| fig-cap: 'Home Ownership'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "darkMode":false, "fontSize":"16px"}}}%%

flowchart TD
    HomeCosts("True Cost<br>of Ownership") --> Mortgage("Mortgage<br>Interest")
    HomeCosts --> PropTax("Property Taxes<br>1-2% annually")
    HomeCosts --> Maintenance("Maintenance<br>~1% annually")
    HomeCosts --> Insurance("Insurance")
    HomeCosts --> Closing("Closing<br>Costs 3-5%")
    HomeCosts --> Opportunity2("Down Payment<br>Opportunity Cost")

    Mortgage --> UnrecovCost("Unrecoverable Costs<br>~5% of home<br>value per year")
    PropTax --> UnrecovCost
    Maintenance --> UnrecovCost
    Insurance --> UnrecovCost
    Closing --> UnrecovCost
    Opportunity2 --> UnrecovCost
    style HomeCosts fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Mortgage fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style PropTax fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Maintenance fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Insurance fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Closing fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Opportunity2 fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style UnrecovCost fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
```

### Felix's "5% Rule" {.unnumbered}

A simplified version of Felix's framework: [the unrecoverable costs of owning are roughly 5% of the home's value per year]{style="color:#f44242; font-weight: bold;"} (property tax ~1%, maintenance ~1%, and cost of capital ~3%, which is the interest and investment return you give up on the money tied up in the house). The math section below walks through the formula and the code.

**Rule of thumb:** Multiply the home's value by 5% and divide by 12. If you can rent an equivalent property for less than that, [renting and investing the difference is likely the better financial choice.]{style="color:#48a56a; font-weight: bold;"}

### When Buying Makes Sense {.unnumbered}

Buying isn't bad, but it should be a [lifestyle decision]{style="color:#48a56a; font-weight: bold;"} more than a financial one. Buy when:

* You're confident you'll stay 7+ years (closing costs need time to amortize)
* You have a stable income and adequate emergency fund
* You value the autonomy, stability, and control
* The local rent-to-price ratio favors buying
* You won't be "house poor" (housing costs should stay under 28-30% of gross income)

### When Renting Makes Sense {.unnumbered}

* You may move within 5-7 years
* Your career requires geographic flexibility
* Local home prices are stretched relative to rents
* You'd rather invest the difference in a diversified portfolio
* You value flexibility, lower maintenance burden, and predictable costs

**The key insight from Felix:** renting is not "throwing money away" any more than buying groceries is throwing money away. [You're paying for shelter either way.]{style="color:#48a56a; font-weight: bold;"} The question is which form of shelter delivers better financial and lifestyle outcomes for your situation.

Putting the pieces together, here is how I'd walk through the decision:

```{mermaid}
%%| fig-cap: 'Rent vs. Buy'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "darkMode":false, "fontSize":"16px"}}}%%

flowchart TD
    Start(["Rent or buy?"]) --> Stay{"Will you stay<br>7+ years?"}
    Stay -->|"No"| RentFlex("Rent and keep<br>your flexibility")
    Stay -->|"Yes"| Rule{"Is rent below the<br>5% rule figure?"}
    Rule -->|"Yes"| RentInvest("Rent and invest<br>the difference")
    Rule -->|"No"| Ready{"Stable income and<br>an emergency fund?"}
    Ready -->|"Yes"| Buy("Buying can make<br>financial sense")
    Ready -->|"No"| Build("Build your<br>cushion first")
    style Start fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Stay fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Rule fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Ready fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style RentFlex fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style RentInvest fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Buy fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Build fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
```

### Mortgage Amortization: The Front-Loaded Interest Reality {.unnumbered}

Mortgages are structured so you pay [mostly interest in early years.]{style="color:#f44242; font-weight: bold;"} Understanding this changes how you think about moving and refinancing.

**Key insight:** On a 30-year mortgage at 6%, more than [80% of your first year's payments go to interest]{style="color:#f44242; font-weight: bold;"}, not principal.

**Rough breakdown of where payments go:**

| Years | Approx. % to Interest | Approx. % to Principal |
|-------|----------------------|------------------------|
| 1-5 | 70-80% | 20-30% |
| 6-15 | 50-70% | 30-50% |
| 16-25 | 25-50% | 50-75% |
| 26-30 | 5-25% | 75-95% |

This is why moving every 5 years is so financially destructive: you're essentially renting from the bank while paying ownership costs on top.

## Math for Major Purchases {#math-for-major-purchases}

Big purchases are where small percentage differences turn into tens of thousands of dollars. Each calculation below is written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter.

### The Rent-vs-Buy 5% Rule {.unnumbered}

The 5% rule isn't magic; it's the sum of three real costs that every homeowner pays whether they realize it or not. The chart splits those costs for a \$500,000 home.

```{r}
#| label: rent-buy-bar-model
#| echo: false
#| fig-cap: "The three components of the 5% rule on a $500,000 home"
#| fig-asp: 0.50
library(ggplot2)
df_5pct <- data.frame(
  component = factor(
    c("Property Tax (~1%)\n~$5,000/yr",
      "Maintenance (~1%)\n~$5,000/yr",
      "Cost of Capital (~3%)\n~$15,000/yr"),
    levels = c(
      "Cost of Capital (~3%)\n~$15,000/yr",
      "Maintenance (~1%)\n~$5,000/yr",
      "Property Tax (~1%)\n~$5,000/yr"
    )
  ),
  amount = c(5000, 5000, 15000)
)
ggplot(df_5pct, aes(x = "", y = amount, fill = component)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = component),
    position = position_stack(vjust = 0.5),
    colour = "#F5F2E8", fontface = "bold", size = 3.5
  ) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Property Tax (~1%)\n~$5,000/yr"    = "#f44242",
    "Maintenance (~1%)\n~$5,000/yr"     = "#0e9aa7",
    "Cost of Capital (~3%)\n~$15,000/yr" = "#8b1a1a"
  )) +
  scale_y_continuous(
    breaks = c(0, 5000, 10000, 15000, 20000, 25000),
    labels = function(x) paste0("$", format(x, big.mark = ","))
  ) +
  labs(x = NULL, y = "Annual unrecoverable cost (Total = $25,000/yr = $2,083/mo)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "none",
    axis.text.y      = element_blank(),
    axis.ticks.y     = element_blank(),
    panel.grid.minor = element_blank()
  )
```

Add the three percentage slices together and you get 5% of home value per year; divide by 12 to get the monthly ownership cost threshold.

**Formula:** Monthly cost of owning ≈ (Home Value × 0.05) ÷ 12

Example: a \$500,000 home → (500,000 × 0.05) ÷ 12 = **\$2,083/month**. Rent below that, and renting-and-investing-the-difference tends to win.

::: panel-tabset
## R

In R, `rent_buy_threshold()` turns the 5% rule into a monthly rent figure.

```{r}
#| label: rent-buy-r
rent_buy_threshold <- function(home_value, unrecoverable_rate = 0.05) {
  (home_value * unrecoverable_rate) / 12
}

# $500,000 home at the standard 5% unrecoverable rate
rent_buy_threshold(home_value = 500000)
```

## Python

The Python version of `rent_buy_threshold()` works the same way:

```{python}
#| label: rent-buy-py
def rent_buy_threshold(home_value, unrecoverable_rate=0.05):
    return (home_value * unrecoverable_rate) / 12

# $500,000 home at the standard 5% unrecoverable rate
rent_buy_threshold(home_value=500000)
```

## Excel

Assuming the home value is in `A2` and the unrecoverable rate (`0.05`) is in `B2`:

``` swift
=(A2*B2)/12
```
:::

### The Monthly Mortgage Payment {.unnumbered}

This is the standard amortization formula behind every mortgage calculator. The same loan produces very different monthly payments depending on how many months you spread it over.

```{r}
#| label: mortgage-payment-bar-model
#| echo: false
#| fig-cap: "Monthly payment on a $480,000 loan at 6% across three terms"
#| fig-asp: 0.50
library(ggplot2)
P <- 480000
r <- 0.06 / 12
terms <- c(15, 20, 30)
n_vals <- terms * 12
payments <- round(P * (r * (1 + r)^n_vals) / ((1 + r)^n_vals - 1))
df_mort <- data.frame(
  term    = factor(paste0(terms, "-Year"), levels = paste0(terms, "-Year")),
  payment = payments
)
ggplot(df_mort, aes(x = term, y = payment, fill = term)) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = paste0("$", format(payment, big.mark = ","))),
    vjust = -0.4, fontface = "bold", size = 4
  ) +
  scale_fill_manual(values = c(
    "15-Year" = "#f44242",
    "20-Year" = "#0e9aa7",
    "30-Year" = "#48a56a"
  )) +
  scale_y_continuous(
    limits = c(0, max(payments) * 1.18),
    labels = function(x) paste0("$", format(x, big.mark = ","))
  ) +
  labs(x = NULL, y = "Monthly payment ($480,000 loan at 6%)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

The formula allocates the principal plus all future interest into equal monthly installments; shorter terms mean higher bars because more principal must be retired each month.

**Formula:** M = P × [ r(1 + r)\^n ] ÷ [ (1 + r)\^n − 1 ]

Where: P = the loan principal, *r* = the monthly interest rate (annual ÷ 12), *n* = the total number of payments (years × 12).

Example: a \$480,000 loan (a \$600,000 home with 20% down) at 6% for 30 years → **\~\$2,878/month**.

::: panel-tabset
## R

In R, `mortgage_payment()` implements the amortization formula.

```{r}
#| label: mortgage-payment-r
mortgage_payment <- function(principal, annual_rate, years) {
  r <- annual_rate / 12
  n <- years * 12
  principal * (r * (1 + r)^n) / ((1 + r)^n - 1)
}

# $480,000 loan at 6% over 30 years
mortgage_payment(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

The Python version of `mortgage_payment()` works the same way:

```{python}
#| label: mortgage-payment-py
def mortgage_payment(principal, annual_rate, years):
    r = annual_rate / 12
    n = years * 12
    return principal * (r * (1 + r) ** n) / ((1 + r) ** n - 1)

# $480,000 loan at 6% over 30 years
mortgage_payment(principal=480000, annual_rate=0.06, years=30)
```

## Excel

Assuming the loan amount is in `A2`, the annual rate is in `B2` (as a decimal), and the years are in `C2`:

``` swift
=A2*((B2/12)*(1+B2/12)^(C2*12))/((1+B2/12)^(C2*12)-1)
```

Excel's built-in `PMT()` function does the same job (it returns a negative number because it treats the payment as money going out, so we flip the sign):

``` swift
=-PMT(B2/12, C2*12, A2)
```
:::

### Where the First Year's Payments Go {.unnumbered}

The front-loaded-interest reality, in numbers: in month one your interest is simply the balance times the monthly rate, and only the rest reduces the principal. Early mortgage payments are mostly interest, which is why moving after a few years means you've barely touched the principal.

```{r}
#| label: first-payment-bar-model
#| echo: false
#| fig-cap: "Interest vs. principal split at three points on a $480,000 / 6% / 30-yr mortgage"
#| fig-asp: 0.50
library(ggplot2)
P <- 480000
r <- 0.06 / 12
n <- 360
pmt <- P * (r * (1 + r)^n) / ((1 + r)^n - 1)
months <- c(1, 180, 348)
balance <- P * (1 + r)^(months - 1) - pmt * ((1 + r)^(months - 1) - 1) / r
interest <- balance * r
principal <- pmt - interest
pct <- function(x) paste0(round(x / pmt * 100), "%")
money <- function(x) paste0("$", format(round(x), big.mark = ","))
df_amort <- data.frame(
  month    = factor(
    rep(c("Month 1\n(Year 1)", "Month 180\n(Year 15)", "Month 348\n(Year 29)"), each = 2),
    levels = c("Month 1\n(Year 1)", "Month 180\n(Year 15)", "Month 348\n(Year 29)")
  ),
  type     = rep(c("Interest", "Principal"), times = 3),
  amount   = as.vector(rbind(interest, principal)),
  label    = as.vector(rbind(
    paste0(money(interest), "\n(", pct(interest), ")"),
    paste0(money(principal), "\n(", pct(principal), ")")
  ))
)
ggplot(df_amort, aes(x = month, y = amount, fill = type)) +
  geom_col(width = 0.55, position = "stack") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    colour = "#F5F2E8", fontface = "bold", size = 3.5
  ) +
  scale_fill_manual(values = c(
    "Interest"  = "#f44242",
    "Principal" = "#48a56a"
  )) +
  scale_y_continuous(
    limits = c(0, pmt * 1.05),
    labels = function(x) paste0("$", format(x, big.mark = ","))
  ) +
  labs(x = NULL, y = paste0("Monthly payment (~", money(pmt), ")"), fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

Each month's interest is just the outstanding balance times the monthly rate; the principal is what's left over after interest is paid.

**Formula:**

* First-month interest = Principal × (annual rate ÷ 12)
* First-month principal = Monthly Payment − First-month interest

Example: on that \$480,000 loan, month-one interest is \$2,400 of the \~\$2,878 payment, about [83% interest]{style="color:#f44242; font-weight: bold;"}, leaving only \~\$478 toward principal.

::: panel-tabset
## R

In R, `first_payment_split()` shows how much of the first payment is interest and how much is principal.

```{r}
#| label: first-payment-r
first_payment_split <- function(principal, annual_rate, years) {
  payment  <- mortgage_payment(principal, annual_rate, years)
  interest <- principal * (annual_rate / 12)
  data.frame(
    payment      = payment,
    interest     = interest,
    principal    = payment - interest,
    pct_interest = (interest / payment) * 100
  )
}

# first payment on a $480,000 loan at 6% over 30 years
first_payment_split(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

The Python version of `first_payment_split()` works the same way:

```{python}
#| label: first-payment-py
def first_payment_split(principal, annual_rate, years):
    payment = mortgage_payment(principal, annual_rate, years)
    interest = principal * (annual_rate / 12)
    return {
        "payment": payment,
        "interest": interest,
        "principal": payment - interest,
        "pct_interest": (interest / payment) * 100,
    }

# first payment on a $480,000 loan at 6% over 30 years
first_payment_split(principal=480000, annual_rate=0.06, years=30)
```

## Excel

Assuming the loan amount is in `A2`, the annual rate is in `B2`, and the years are in `C2`, the monthly payment goes in `D2`:

``` swift
=-PMT(B2/12, C2*12, A2)
```

The first month's interest goes in `E2`:

``` swift
=A2*B2/12
```

The principal portion goes in `F2`:

``` swift
=D2-E2
```

And the percent of the payment that is interest goes in `G2` (format the cell as a **Percentage**):

``` swift
=E2/D2
```
:::

## Recap

1. **Renting isn't throwing money away.** You pay for shelter either way. The real question is which option leaves you better off for your situation.
2. **Use the 5% rule as a quick test.** Multiply the home's value by 5% and divide by 12. On a \$500,000 home that's about \$2,083 a month. If you can rent a comparable place for less, renting and investing the difference usually wins.
3. **Owning costs more than the mortgage.** Property tax, maintenance, and the return you give up on your down payment add up to roughly 5% of the home's value each year.
4. **Early mortgage payments are mostly interest.** In the first year, more than 80% of each payment goes to interest, which is why moving after a few years leaves you with barely any principal paid down.
5. **Buy for lifestyle reasons, not just math.** Buy when you plan to stay 7 or more years, your income is stable, you have an emergency fund, and your housing costs stay under 28 to 30% of your gross income.

A house is the biggest purchase most people ever make, so run the numbers before you fall in love with a floor plan.
