# Managing Debt

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

Not all debt is equal. A 3% mortgage and a 24% credit card balance call for very different strategies.

  - Good debt vs. bad debt    
  - The **avalanche** method (highest rate first) vs. the **snowball** method (smallest balance first)    
  - Refinancing and consolidation   
  - When to prioritize debt payoff over investing   
  
Debt isn't inherently good or bad; it's a tool. Used wisely, it can accelerate wealth-building (think low-rate mortgages or strategic education). Used poorly, it can quietly destroy decades of financial progress. Let's break down each type with evidence-based advice from Housel, Sethi, Bogle, and Ben Felix.

## The Debt Hierarchy: Know Your Enemy

Not all debt is created equal. Attack it in order of damage potential.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Know Your Enemy'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    AllDebt[Your Debts] --> Toxic[TOXIC DEBT<br>15-29% APR]
    AllDebt --> Moderate[MODERATE DEBT<br>5-10% APR]
    AllDebt --> Strategic[STRATEGIC DEBT<br>3-7% APR]
    
    Toxic --> CreditCards[Credit Cards<br>Payday Loans]
    Moderate --> CarLoans[Car Loans<br>Private Student Loans]
    Strategic --> Mortgages[Mortgages<br>Federal Student Loans]
    
    CreditCards --> Attack1[Attack Aggressively<br>Pay Off First]
    CarLoans --> Attack2[Pay Down Steadily<br>Avoid New Ones]
    Mortgages --> Attack3[Pay on Schedule<br>Invest the Difference]
```

## Credit Cards: The Wealth Destroyer

Credit card debt is the single most destructive financial product most people will ever encounter. At 20%+ APR, it grows faster than any reasonable investment can compound.

**Ramit Sethi's approach is uncompromising here:** credit card debt must be eliminated before any other financial goal beyond a starter emergency fund. The math is brutal: you cannot out-invest 24% interest.

### The Playbook {.unnumbered}

| Step | Action |
|------|--------|
| **1. Stop the bleeding** | Stop using cards. Pay cash or debit until balances are zero. |
| **2. Call and negotiate** | Call each issuer. Ask for a lower APR. It works more often than you'd think. |
| **3. Choose a method** | **Avalanche** (highest APR first; saves most money) or **Snowball** (smallest balance first; builds momentum). |
| **4. Consider consolidation** | A 0% balance transfer card or personal loan at 8-10% beats 24% every time. |
| **5. Automate minimums** | Never miss a payment; it triggers penalty APRs and credit damage. |

### Use Credit Cards as a Tool, Not a Trap {.unnumbered}

Once you're paying off your balance **in full every month**, credit cards become powerful: cashback, travel rewards, fraud protection, and credit-building. The key is using them as a payment method, not a financing tool.

## Student Loans: A Nuanced Approach

Student loans sit in a complicated middle ground. Ben Felix's human capital framework applies powerfully here: education is an investment in your future earning potential, but only when the math works.

### Before Taking Loans {.unnumbered}

Felix would frame it this way: **the return on educational debt depends on the income premium that degree generates.** A \$40,000 loan for a degree that boosts lifetime earnings by $1M+ is a great trade. The same \$40,000 for a degree with no clear earnings path is not.

### After Taking Loans {.unnumbered}

```{mermaid}
%%| fig-cap: ' Student Loans'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    StudentLoans[Student<br>Loan<br>Balance] --> Type{Federal<br>or Private?}
    Type -->|Federal| FedOptions[Federal<br>Options]
    Type -->|Private| PrivOptions[Private<br>Options]
    
    FedOptions --> IDR[Income-Driven<br>Repayment]
    FedOptions --> PSLF[Public<br>Service<br>Forgiveness]
    FedOptions --> Standard[Standard<br>10-Year]
    
    PrivOptions --> Refi[Refinance<br>for<br>Lower Rate]
    PrivOptions --> PayDown[Aggressive<br>Paydown]
```

### The Pay-Off vs. Invest Decision {.unnumbered}

A key insight from Felix's research-driven approach: **compare your loan interest rate to your expected investment return after tax.**

- **Interest rate above ~6%:** Prioritize paying off    
- **Interest rate 4-6%:** Roughly a wash; do both
- **Interest rate below 4%:** Pay the minimum, invest the difference    

Don't forget the **psychological factor** Housel emphasizes: some people sleep better debt-free, even if the math slightly favors investing. That peace of mind has real value.

## Car Payments: The Wealth Killer in Disguise

Cars are where middle-class wealth quietly dies. Here's why:

```{mermaid}
%%| fig-cap: 'Car Payments'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    NewCar[New Car<br>Purchase] --> Depreciation[Loses 20-30%<br>in Year 1]
    NewCar --> Payment[$500-700+<br>Monthly Payment]
    NewCar --> Insurance[Higher<br>Insurance]
    NewCar --> Opportunity[Opportunity<br>Cost]
    Opportunity --> LostWealth[$600/mo<br>invested over<br>30 years<br>= ~$700,000]
```

### Ben Felix's Perspective {.unnumbered}

Felix has been blunt: **cars are depreciating consumption assets, not investments.** Financing a depreciating asset at 7-9% interest while it loses value is one of the worst financial moves available, yet it's culturally normalized.

### The Rules of Smart Car Ownership {.unnumbered}

1. **Buy used, 2-4 years old.** Let someone else absorb the steep first-year depreciation.
2. **Pay cash when possible.** If you can't pay cash, you can't really afford it.
3. **If financing, keep it under 36 months** at the lowest rate available.
4. **The 20/4/10 rule:** 20% down, financed no longer than 4 years, total transportation costs under 10% of gross income.
5. **Drive it until the wheels fall off.** The cheapest car you'll ever own is the reliable one you already have.

**The status trap:** Housel's insight applies powerfully here. Nobody is impressed by your car; they're imagining themselves in it. You're paying a fortune for a status signal no one is sending back.

## Putting It All Together

```{mermaid}
%%| fig-cap: 'Put It All Together'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Start[Manage Your Debt] --> Step1["Eliminate Credit<br>Card Debt"]
    Step1 --> Step2["Build Emergency Fund<br>3-6 Months Expenses"]
    Step2 --> Step3["Address Student Loans<br>Based on Interest Rate"]
    Step3 --> Step4["Drive Used,<br>Modest Cars"]
    Step4 --> Freedom["Financial Freedom"]
```

## Key takeaways

Debt management isn't about avoiding all debt; it's about being **intentional** with it. The right framework:

1. **Eliminate toxic debt ruthlessly.** Credit cards first, always.
2. **Think in human capital.** Student loans are an investment; evaluate the return.
3. **Resist depreciating debt.** Cars financed at high rates are wealth-killers.

As Housel reminds us: **the goal isn't to be debt-free for its own sake. The goal is freedom.** And freedom comes from controlling your money before it controls you.

## Math for Managing Debt

Debt is fundamentally a math problem dressed up in emotional clothing. Lenders count on you not running the numbers, because if you did, many borrowing decisions would look very different. Here are the calculations every borrower should be fluent in.

Each calculation is also written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter, now applied to lending math.

### APR vs. APY: Know What You're Actually Paying {.unnumbered}

These two terms get used interchangeably, but they're different, and lenders often quote whichever number makes the loan look better.

**APR (Annual Percentage Rate):** The stated annual interest rate, not accounting for compounding.

**APY (Annual Percentage Yield):** The effective annual rate, including the impact of compounding.

**Formula:** APY = (1 + APR/n)^n − 1

Where *n* = number of compounding periods per year.

Example: A credit card with a 24% APR compounded daily:
APY = (1 + 0.24/365)^365 − 1 = **~27.1%**

That 3.1% gap is real money. On a $10,000 balance, it's an extra $310/year you didn't realize you were paying.

::: {.panel-tabset}

## R

```{r}
#| label: apr-apy-r
apr_to_apy <- function(apr, periods_per_year = 365) {
  (1 + apr / periods_per_year)^periods_per_year - 1
}

# 24% APR compounded daily
apr_to_apy(apr = 0.24)
```

## Python

```{python}
#| label: apr-apy-py
def apr_to_apy(apr, periods_per_year=365):
    return (1 + apr / periods_per_year) ** periods_per_year - 1

# 24% APR compounded daily
apr_to_apy(apr=0.24)
```

:::

### The True Cost of Minimum Payments {.unnumbered}

This is the single most important debt calculation you can master. Credit card companies set minimum payments low because **the longer you pay, the more they earn.**

**Quick approximation for credit card payoff time:**

> Months to pay off ≈ Balance ÷ (Monthly Payment − Monthly Interest)

Where: Monthly Interest = Balance × (APR ÷ 12)

Example: $5,000 balance at 22% APR, paying $100/month minimum:
- Monthly interest: $5,000 × (0.22/12) = ~$92
- Effective principal reduction: $100 − $92 = **$8/month**
- Payoff time: $5,000 ÷ $8 = **~625 months (52 years)**

That's not a typo. Minimum payments are designed to keep you in debt for life.

::: {.panel-tabset}

## R

```{r}
#| label: min-payoff-r
months_to_payoff <- function(balance, monthly_payment, apr) {
  monthly_interest    <- balance * (apr / 12)
  principal_reduction <- monthly_payment - monthly_interest
  balance / principal_reduction
}

# $5,000 balance, $100/month, 22% APR
months_to_payoff(balance = 5000, monthly_payment = 100, apr = 0.22)
```

## Python

```{python}
#| label: min-payoff-py
def months_to_payoff(balance, monthly_payment, apr):
    monthly_interest    = balance * (apr / 12)
    principal_reduction = monthly_payment - monthly_interest
    return balance / principal_reduction

# $5,000 balance, $100/month, 22% APR
months_to_payoff(balance=5000, monthly_payment=100, apr=0.22)
```

:::

### Total Interest Paid Over a Loan's Life {.unnumbered}

For any installment loan (car, mortgage, student loan), this is the number that should drive your decisions.

**Formula:** Total Interest = (Monthly Payment × Number of Payments) − Loan Principal

Example: $30,000 car loan at 7% for 6 years (~$512/month):
- Total paid: $512 × 72 = $36,864
- Total interest: $36,864 − $30,000 = **$6,864 in interest**

Now run the same loan at 4 years instead of 6:
- Monthly payment: ~$719
- Total paid: $719 × 48 = $34,512
- Total interest: **$4,512**

You save **$2,352** by shortening the term: the math behind the 20/4/10 car rule.

::: {.panel-tabset}

## R

```{r}
#| label: total-interest-r
total_interest <- function(monthly_payment, n_months, principal) {
  (monthly_payment * n_months) - principal
}

# 6-year car loan vs. 4-year, $30,000 principal
total_interest(monthly_payment = 512, n_months = 72, principal = 30000)
total_interest(monthly_payment = 719, n_months = 48, principal = 30000)
```

## Python

```{python}
#| label: total-interest-py
def total_interest(monthly_payment, n_months, principal):
    return (monthly_payment * n_months) - principal

# 6-year car loan vs. 4-year, $30,000 principal
print(total_interest(monthly_payment=512, n_months=72, principal=30000))
print(total_interest(monthly_payment=719, n_months=48, principal=30000))
```

:::

### Debt-to-Income Ratio (DTI) {.unnumbered}

This is what lenders calculate about you. You should calculate it about yourself.

**Formula:** DTI = (Total Monthly Debt Payments ÷ Gross Monthly Income) × 100

Example: $2,500 in monthly debt payments on $7,000 gross income → 2,500 ÷ 7,000 = **~36% DTI**

Benchmarks:
- **Below 20%:** Healthy
- **20-35%:** Manageable but watch carefully
- **36-43%:** Stretched; lenders get cautious
- **Above 43%:** Danger zone; most mortgage lenders won't approve you

::: {.panel-tabset}

## R

```{r}
#| label: dti-r
dti_ratio <- function(monthly_debt_payments, gross_monthly_income) {
  (monthly_debt_payments / gross_monthly_income) * 100
}

# $2,500 in debt payments on $7,000 gross income
dti_ratio(monthly_debt_payments = 2500, gross_monthly_income = 7000)
```

## Python

```{python}
#| label: dti-py
def dti_ratio(monthly_debt_payments, gross_monthly_income):
    return (monthly_debt_payments / gross_monthly_income) * 100

# $2,500 in debt payments on $7,000 gross income
dti_ratio(monthly_debt_payments=2500, gross_monthly_income=7000)
```

:::

### The Pay-Off vs. Invest Comparison {.unnumbered}

This is the math behind every "should I pay extra on my mortgage or invest?" debate.

**The rule of thumb:**

> If after-tax loan interest rate > expected after-tax investment return → Pay off the debt
> If after-tax loan interest rate < expected after-tax investment return → Invest the difference

Example: 5% mortgage vs. expected 7% stock market return (real):
- The math favors investing by ~2% per year
- **But:** the loan return is **guaranteed**, the investment return is not

This is where Housel's wisdom matters: **a guaranteed 5% beats a probable 7% for many people psychologically.** Run the math, then add your peace-of-mind premium.

::: {.panel-tabset}

## R

```{r}
#| label: payoff-vs-invest-r
payoff_vs_invest <- function(loan_rate, expected_return) {
  if (loan_rate > expected_return) {
    "Pay off the debt"
  } else if (loan_rate < expected_return) {
    "Invest the difference"
  } else {
    "Roughly a wash; do both"
  }
}

# 5% mortgage vs. expected 7% return
payoff_vs_invest(loan_rate = 0.05, expected_return = 0.07)
```

## Python

```{python}
#| label: payoff-vs-invest-py
def payoff_vs_invest(loan_rate, expected_return):
    if loan_rate > expected_return:
        return "Pay off the debt"
    elif loan_rate < expected_return:
        return "Invest the difference"
    else:
        return "Roughly a wash - do both"

# 5% mortgage vs. expected 7% return
payoff_vs_invest(loan_rate=0.05, expected_return=0.07)
```

:::

### The Avalanche vs. Snowball Math {.unnumbered}

For multiple debts, the choice between methods has a calculable cost.

**Avalanche method:** Highest APR first. Mathematically optimal.
**Snowball method:** Smallest balance first. Psychologically motivating.

**The hidden cost of snowball:**

To calculate, find total interest paid under each method using a debt payoff calculator. The difference is what you're "paying" for the psychological win of quick early payoffs.

For most people with moderate debt, the snowball costs $200-$1,000 extra in interest. **If that's the price of actually finishing the journey, it's often worth it.** If you're disciplined enough for avalanche, take the savings.

These two functions take a list of your debts and return them in the order each method says to attack them.

::: {.panel-tabset}

## R

```{r}
#| label: avalanche-snowball-r
# avalanche: highest APR first
order_avalanche <- function(debts) {
  debts[order(-debts$apr), ]
}

# snowball: smallest balance first
order_snowball <- function(debts) {
  debts[order(debts$balance), ]
}

# your debts, as a small table
debts <- data.frame(
  name    = c("Card A", "Card B", "Card C"),
  balance = c(1000, 5000, 2000),
  apr     = c(0.18, 0.24, 0.15)
)

order_avalanche(debts)
order_snowball(debts)
```

## Python

```{python}
#| label: avalanche-snowball-py
# avalanche: highest APR first
def order_avalanche(debts):
    return sorted(debts, key=lambda d: d["apr"], reverse=True)

# snowball: smallest balance first
def order_snowball(debts):
    return sorted(debts, key=lambda d: d["balance"])

# your debts, as a list of small records
debts = [
    {"name": "Card A", "balance": 1000, "apr": 0.18},
    {"name": "Card B", "balance": 5000, "apr": 0.24},
    {"name": "Card C", "balance": 2000, "apr": 0.15},
]

order_avalanche(debts)
order_snowball(debts)
```

:::

### The Credit Utilization Ratio {.unnumbered}

For credit score health, this calculation matters month to month.

**Formula:** Utilization = (Current Balance ÷ Credit Limit) × 100

Keep this under **30%** on any individual card and across all cards in total. Under **10%** is ideal for maximum score impact.

Example: $3,000 balance on $10,000 limit = **30% utilization**, right at the threshold.

::: {.panel-tabset}

## R

```{r}
#| label: utilization-r
credit_utilization <- function(balance, credit_limit) {
  (balance / credit_limit) * 100
}

# $3,000 balance on a $10,000 limit
credit_utilization(balance = 3000, credit_limit = 10000)
```

## Python

```{python}
#| label: utilization-py
def credit_utilization(balance, credit_limit):
    return (balance / credit_limit) * 100

# $3,000 balance on a $10,000 limit
credit_utilization(balance=3000, credit_limit=10000)
```

:::

## Putting the Math to Work

```{mermaid}
%%| fig-cap: 'Putting the Math to Work'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    DebtMath[Debt Math Mastery] --> Decision1[Before Borrowing:<br/>Calculate Total Interest]
    DebtMath --> Decision2[While Paying Off:<br/>Compare APR to Returns]
    DebtMath --> Decision3[Managing Credit:<br/>Monitor DTI & Utilization]

    Decision1 --> Outcome[Intentional<br/>Borrowing Decisions]
    Decision2 --> Outcome
    Decision3 --> Outcome

    Outcome --> Freedom[Financial Freedom]
```

## Key takeaways

You don't need to be a math genius to manage debt well, but you do need to be **numerically fluent** in these specific calculations. Why? Because lenders absolutely are.

The math always tells the truth. Marketing tells you what's possible. Emotions tell you what feels right. **The math tells you what's actually happening to your money.**

As Bogle reminded investors: "The miracle of compounding returns is overwhelmed by the tyranny of compounding costs." The same principle applies in reverse to debt: **compounding interest is wealth-destroying when it's working against you.** Master the math, and you take that weapon out of the lender's hands.