# R Reference {.unnumbered}

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

A quick reference to the R concepts used throughout this book.

## Functions in R

Functions perform operations (calculate, transform, model, graph) on various objects that contain information (blood pressure measurements, monthly sales, political party affiliation, etc.)

```{r}
#| label: r-function
percent_of_income <- function(expense, income) {
  (expense / income) * 100
}

percent_of_income(expense = 1500, income = 5000)
```

The pattern is `name <- function(inputs) { body }`. The last expression in the body is returned automatically; `return()` is optional.

## Types of objects in R

### Vectors

Vectors are one-dimensional containers where every value has the same type.

#### Atomic vectors {.unnumbered}

The four atomic types: **logical**, **integer**, **double**, and **character**.

```{r}
#| label: r-atomic
c(TRUE, FALSE, TRUE)              # logical
c(1L, 2L, 3L)                     # integer
c(1.5, 2.0, 3.14)                 # double
c("apple", "banana", "cherry")    # character
```

#### S3 vectors {.unnumbered}

S3 vectors (**factors**, **dates**, **date-times**, **durations**) are atomic vectors with extra class metadata that gives them specialized behavior.

```{r}
#| label: r-s3
factor(c("low", "med", "high"),
       levels = c("low", "med", "high"))   # factor
as.Date("2026-01-15")                      # date
as.POSIXct("2026-01-15 09:30:00")          # date-time
as.difftime(2, units = "hours")            # duration
```

### Matrices

Two dimensional objects. All values must share the same atomic type.

```{r}
#| label: r-matrix
matrix(1:6, nrow = 2, ncol = 3)
```

### Arrays

Multidimensional objects. Same-type values; dimensions defined by `dim()`.

```{r}
#| label: r-array
array(1:12, dim = c(2, 3, 2))
```

### Data frames & tibbles

Rectangular objects: each column is a vector, and columns can be different types. Tibbles are tidyverse-style data frames with nicer printing and stricter behavior.

```{r}
#| label: r-tibble
tibble::tibble(
  name = c("Alice", "Bob", "Carol"),
  age  = c(34L, 28L, 41L),
  paid = c(TRUE, FALSE, TRUE)
)
```

### Lists

Recursive objects: each element can be any object, including another list. The most flexible container in R, and the building block underneath data frames, model objects, and JSON-like data.

```{r}
#| label: r-list
list(
  numbers = c(1, 2, 3),
  name    = "Alice",
  flags   = c(TRUE, FALSE),
  nested  = list(a = 1, b = "two")
)
```
