# Savings Strategies

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

How much should you save? The honest answer is: **as much as you can
sustain**. This chapter covers common targets and how to choose one that
matches your goals.

  - Pay-yourself-first
  - Sinking funds for known future expenses
  - The savings rate as the single most important number

## Math for Saving {#math-for-saving}

### Your Savings Rate (Your Most Important Metric) {.unnumbered} 

[Ben Felix](https://rationalreminder.ca/) and [John Bogle](https://www.bogleheads.org/) both stress this above all else.

**Formula:** (Savings + Investments) ÷ Gross Income

Example: You save $500 and invest $700 from a $5,000 gross income. → 1,200 ÷ 5,000 = **24% savings rate.**

Benchmarks to know:

- **<10%:** Danger zone; build aggressively
- **10-20%:** Average; you'll retire eventually
- **20-30%:** Strong; you're building real wealth
- **30%+:** Path to early financial independence  

Your savings rate, the share of take-home pay you *don't* spend, is the single number that best predicts how quickly you build wealth. It's Housel's "gap between ego and income," expressed as a percentage.

**Formula:** Savings Rate = (Income − Spending) ÷ Income × 100

Example: you take home \$5,000 and spend \$3,500. → (5,000 − 3,500) ÷ 5,000 = **30%**.

::: panel-tabset
## R

```{r}
#| label: savings-rate-r
savings_rate <- function(income, spending) {
  ((income - spending) / income) * 100
}

# $5,000 take-home, $3,500 spent
savings_rate(income = 5000, spending = 3500)
```

## Python

```{python}
#| label: savings-rate-py
def savings_rate(income, spending):
    return ((income - spending) / income) * 100

# $5,000 take-home, $3,500 spent
savings_rate(income=5000, spending=3500)
```
:::