# Tracking Spending

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

<br>

You only manage what you measure. This chapter covers how to capture every dollar that leaves your accounts and turn that data into insight.

  - Capturing every transaction
  - Working with digital bank exports
  - Categorizing what you've captured
  - Reviewing on a monthly cadence
  - Spotting "leak" categories (subscriptions, delivery fees, impulse buys)

Tracking is the *recording* side of personal finance. [Budgeting](budgeting.qmd) covers the *allocation* side — where money is supposed to go. Without tracking, a budget is a guess; with it, the budget becomes a feedback loop.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Budget Mindset'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Budgeting(["<strong>BUDGETING</strong><br/>Where Money<br>Should Go"]) --> Plan("<em>Plan</em>")
    Tracking(["<strong>TRACKING</strong><br/>Where Money<br>Actually Went"]) --> Reality("<em>Reality</em>")

    Plan --> Compare{"<strong>COMPARE</strong>:<br/>Plan vs. Reality"}
    Reality --> Compare

    Compare --> Insights("Insights &<br>Adjustments")
    Insights --> FeedbackLoop(["<em>Feedback</em>"])
    FeedbackLoop --> Budgeting
    style Budgeting fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Tracking fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Plan fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Reality fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Compare fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Insights fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style FeedbackLoop fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
```


## Capturing Transactions

below are a few time-tested ways to capture your spending. If you're reading this, you'll also know there's a sea of apps and digital tools for tracking your spending. It's important to know people were tracking their spending before the internet, personal computers, and calculators. Regardless of the method you choose, the important thing is to pick one, run it for a full month, then decide whether if it works for you. 

### A spending journal {.unnumbered}

A small notebook, kept in a pocket or bag. Every time money leaves your hands you write one line: amount, place, category. At week's end, you tally.

This is the oldest method on the list — Benjamin Franklin kept one. Its strength is the moment of awareness it forces. Writing "$6 — Starbucks — coffee" by hand is a small pause that an automatically-recorded card swipe will never give you.

### Checkbook registers {.unnumbered}

Every paper checkbook used to come with a register: each transaction logged on its own line, with a running balance. It works for any account where you can list debits and credits. One page per month is plenty. The discipline isn't the form — it's that writing each entry keeps you honest about the balance.

### Cash envelope system {.unnumbered}

For variable categories (groceries, dining, gas, fun money), withdraw the monthly amount in cash and split it into physical envelopes labeled by category. When an envelope is empty, that category is done for the month. No swiping, no spreadsheet to update.

Depression-era families used envelopes because they had to. The system has stuck around because the physical, finite nature of cash slows spending more reliably than any reminder ever will.

### Bank statements {.unnumbered}

If you live almost entirely on debit and credit cards, every transaction is already captured for you. The work shifts from capture to **review**: once a month, pull the bank and card statements, go through them line by line, and assign each transaction to a category. This is the modern equivalent of the ledger, except the bank has done the writing.

## Using Spreadsheets

Digital tracking is the bank-statement method with the math done for you. A real export from your bank includes the same four columns as the example below: a **date**, a **description**, a **category**, and an **amount** (negative for outgoing money, positive for incoming).

::: panel-tabset
## R

Create the `bank_statement` data in R:

```{r}
#| label: bank-statement-r
#| code-fold: true
#| code-summary: 'show/hide R bank_statement'
bank_statement <- tibble::tibble(
  Date = as.Date(c(
    "2026-01-02", "2026-01-02", "2026-01-03", "2026-01-03",
    "2026-01-04", "2026-01-04", "2026-01-05", "2026-01-06",
    "2026-01-07", "2026-01-08", "2026-01-09", "2026-01-10",
    "2026-01-11", "2026-01-12", "2026-01-13", "2026-01-14",
    "2026-01-15"
  )),
  Description = c(
    "Employer Payroll",       "Rent",
    "Comcast Internet",       "Trader Joe's",
    "Shell Gas",              "Netflix",
    "Starbucks",              "PG&E Electric",
    "AT&T Wireless",          "Spotify",
    "Whole Foods",            "Chipotle",
    "Geico Auto Insurance",   "NYTimes Digital",
    "Shell Gas",              "Costco",
    "24 Hour Fitness"
  ),
  Category = c(
    "Income",         "Housing",
    "Utilities",      "Groceries",
    "Transportation", "Subscription",
    "Dining",         "Utilities",
    "Utilities",      "Subscription",
    "Groceries",      "Dining",
    "Insurance",      "Subscription",
    "Transportation", "Groceries",
    "Subscription"
  ),
  Amount = c(
     2400.00, -1500.00,
      -75.00,   -82.45,
      -45.20,   -15.99,
       -6.75,  -142.30,
      -85.00,   -10.99,
      -64.12,   -14.50,
     -125.00,   -17.00,
      -42.80,  -156.78,
      -39.99
  )
)
```

```{r}
#| label: print-bank-statement-r
#| df_print: tibble
bank_statement
```

## Python

Create the `bank_statement` data in Python (with `pandas`):

```{python}
#| label: bank-statement-py
#| code-fold: true
#| code-summary: 'show/hide Python bank_statement'
import pandas as pd

bank_statement = pd.DataFrame({
    "Date": pd.to_datetime([
        "2026-01-02", "2026-01-02", "2026-01-03", "2026-01-03",
        "2026-01-04", "2026-01-04", "2026-01-05", "2026-01-06",
        "2026-01-07", "2026-01-08", "2026-01-09", "2026-01-10",
        "2026-01-11", "2026-01-12", "2026-01-13", "2026-01-14",
        "2026-01-15"
    ]),
    "Description": [
        "Employer Payroll",     "Rent",
        "Comcast Internet",     "Trader Joe's",
        "Shell Gas",            "Netflix",
        "Starbucks",            "PG&E Electric",
        "AT&T Wireless",        "Spotify",
        "Whole Foods",          "Chipotle",
        "Geico Auto Insurance", "NYTimes Digital",
        "Shell Gas",            "Costco",
        "24 Hour Fitness"
    ],
    "Category": [
        "Income",         "Housing",
        "Utilities",      "Groceries",
        "Transportation", "Subscription",
        "Dining",         "Utilities",
        "Utilities",      "Subscription",
        "Groceries",      "Dining",
        "Insurance",      "Subscription",
        "Transportation", "Groceries",
        "Subscription"
    ],
    "Amount": [
         2400.00, -1500.00,
          -75.00,   -82.45,
          -45.20,   -15.99,
           -6.75,  -142.30,
          -85.00,   -10.99,
          -64.12,   -14.50,
         -125.00,   -17.00,
          -42.80,  -156.78,
          -39.99
    ]
})
```

```{python}
#| label: print-bank-statement-py
bank_statement
```
:::

Once your transactions sit in columns and rows, it's much easier to answer tracking questions like:

1. What is the total spent on subscriptions?
2. What percentage of my income is going to fixed costs?
3. What was my biggest variable category last month?

Bogle would approve: **simplicity wins.** A well-built spreadsheet beats most apps because it forces you to engage with your numbers.

## Categorizing Transactions

Most people overcomplicate categories. Six to ten buckets is plenty, and a small, stable list will outlast any rotating set of "smart" categorizations. A working starter set:

- **Housing** (rent/mortgage, utilities, insurance)
- **Transportation** (car payment, fuel, maintenance, transit)
- **Food — groceries**
- **Food — dining out**
- **Subscriptions** (everything recurring and digital)
- **Personal** (clothes, haircuts, gym)
- **Health & insurance**
- **Gifts & giving**
- **Fun money** (don't break this down further — protect it from over-tracking)
- **Other / one-offs**

Two rules that save hours of bookkeeping:

1. **When in doubt, "Other".** Don't invent a new category for a single transaction. Review the "Other" bucket at month-end and only promote a new category once you've seen the same kind of transaction three months in a row.
2. **Keep the categories stable.** Changing categories every month destroys the comparisons that make tracking useful in the first place.

The `bank_statement` dataset defined above is already organized this way: each row carries a `Category` alongside its `Amount`, so totalling by category becomes a single grouping operation.

## The Monthly Review

Tracking is only valuable if you actually look at the totals. A monthly review is a thirty-minute habit, scheduled on the same day each month — the day after your statements close is a natural anchor.

The ritual has four steps:

1. **Pull** the bank and credit-card statements for the month.
2. **Categorize** every line into your buckets, using whichever capture method you've chosen.
3. **Compare** category totals to your budget. Where did you go over? Where did you come in under?
4. **Pick one change** for next month. *One.* The point is sustainable adjustment, not perfection.

Over a year, twelve small adjustments compound into a budget that actually fits your life. That's the real return on tracking: the calibration loop, not the spreadsheet itself.

## Online Subscriptions

This is the most overlooked category in personal finance. Streaming, apps, software, memberships, deliveries; they're individually small but collectively enormous and can silently drain your accounts. 

```{mermaid}
%%| fig-cap: 'Subscription Audit'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%

flowchart TD
    Subscriptions[Monthly Subscriptions] --> Audit[Quarterly Audit]
    Audit --> Use[Used in last 30 days?]
    Use -->|Yes| Worth[Worth the cost?]
    Use -->|No| Cancel[Cancel Today]
    Worth -->|Yes| Keep[Keep]
    Worth -->|No| Cancel
    Cancel --> Redirect[Redirect to Investments]
    style Subscriptions fill:#48a56a,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Audit fill:#0e9aa7,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Use fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Worth fill:#f0cfcf,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Cancel fill:#f44242,color:#F5F2E8,stroke:#1C1C1E,stroke-width:2px
    style Keep fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
    style Redirect fill:#86ddcd,color:#1C1C1E,stroke:#1C1C1E,stroke-width:2px
```

### The Subscription Reality Check {.unnumbered}

The average household spends **$200–$300/month** on subscriptions, often without realizing it. At $250/month invested at a 7% real return over 30 years, that's roughly **$300,000 in lost wealth**.

**Action steps:**

1. Pull your last 3 months of bank/card statements
2. List every recurring charge
3. Cancel anything you haven't actively used in 30 days
4. Set a calendar reminder every 90 days to repeat this
5. Use one card exclusively for subscriptions to track them easily

This isn't deprivation; it's the Sethi principle in action: **cut mercilessly on what you don't love** so you can spend freely on what you do.

## Math for tracking spending

Below are some tips and tools for tracking spending. Some can be used to easily categorize purchases, others can be used to check the amount you're spending on subscriptions. 

### The Subscription Audit Math {.unnumbered}

A simple but powerful calculation.

**Formula:** Monthly Subscription × 12 × (1.07)^n = Opportunity cost in *n* years

Example: A $20/month subscription you don't really use, over 30 years:

- Annual cost: $240
- Invested at 7% real return for 30 years: **~$24,000**

That "small" $20/month is a $24,000 decision. **Audit accordingly.**
